---
title: "ChiBlock Builder Data Exploration Report"
author: 
  - "**Great Cities Institute**"
  - "Kyle McFarren"
date: today
editor: 
  mode: visual
  markdown: 
    wrap: sentence
format:
  html:
    echo: false
    theme: sandstone
    toc: true
    toc_float: true
    number_sections: true
  pdf:
    number_sections: true
execute:
  eval: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(leaflet)
library(sf)
library(tigris)
library(forcats)
library(knitr)
library(tidyr)
library(tmap)
library(ggmap)
#library(leaflet.extras)
library(janitor)

options(dplyr.summarise.inform = FALSE)

#hiding these options here...
#code-fold: true
#    code-summary: "Show the code"

setwd("C:/GCI/CBB R")

knitr::opts_knit$set(root.dir = "C:/GCI/CBB R")
#knitr::opts_knit$set(root.dir = "C:/Users/kylem/MSCA/GCI/BlockBuilder Data/Intro Report")
data_clean <- read.csv("CBB_Data_Rebuild3.3.csv")

data_clean <- data_clean %>% select(1:37)

data_clean <- data_clean %>%
  mutate(Property.PIN = as.character(Property.PIN))

#data_clean <- read.csv("CBB_Data_Rebuild3.3.csv")
neighborhood_key <- read.csv("Neighborhood connection key.csv")
neighborhood_key <- neighborhood_key %>%
  dplyr::select(1, 2)

neighborhood_key <- neighborhood_key %>%
  rename(connection_time = `Length.of.time.that.applicant.has.had.connection.to.neighborhood`)

data_clean <- data_clean %>%
  left_join(neighborhood_key, by = c("connection" = "connection_time"))

#data_clean %>% filter(is.na(Translate.x))
#227 are NA

data_clean %>% filter(connection == "")
#267 are from a blank connection, those are ok

#data_clean %>% filter(connection != "") %>%
#  filter(is.na(Translate.x)) %>%
#  dplyr::select(connection)

#0 lines, all responses accounted for now

# data_clean <- data_clean %>%
#   rename(connection = connection, connection6 = Translate.x, applicant_ZIP = applicant_ZIP, hispanic = hispanic, LGBTQIA = LGBTQIA, age = age, income = income) %>%
#   dplyr::select(Record.ID, Name.Key, Person.Count, applicant_ZIP, Application.Status, Application.Type, Property.PIN, Property.Applications, Date.Submitted, Property.Type.of.Land, Property.Address, connection2, Applicant.Race, hispanic, Gender.Identity, LGBTQIA,  everything())

# Clean up the entity types in the data
data_clean <- data_clean %>%
  mutate(Purchasing.Entity.Type = case_when(
    Purchasing.Entity.Type %in% c("For-Profit Business Corporation", "For Profit Busines Corporation", "For-profit corporation") ~ "For-Profit Organization",
    Purchasing.Entity.Type == "Individual" ~ "Sole Proprietor/Individual",
    is.na(Purchasing.Entity.Type) | Purchasing.Entity.Type == "" ~ "Not Provided",
    TRUE ~ Purchasing.Entity.Type  # Leave other types as is
  ))

# Clean up any invalid UTF-8 characters in Name.Key and Purchasing.Entity.Type
data_clean <- data_clean %>%
  mutate(Name.Key = iconv(Name.Key, from = "UTF-8", to = "ASCII", sub = ""),
         Purchasing.Entity.Type = iconv(Purchasing.Entity.Type, from = "UTF-8", to = "ASCII", sub = ""))

data_clean %>%
  count(Application.Status)

data_clean$Application.Status2 <- data_clean$Application.Status  # Step 1: Create the field

# Step 2: Recode "Withdrawn" or "Declined" into "Declined & Withdrawn"
data_clean$Application.Status2 <- ifelse(data_clean$Application.Status %in% c("Withdrawn", "Declined"),
                                         "Declined & Withdrawn",
                                         data_clean$Application.Status2)

# Replace blanks/NA in Purchasing.Entity.Type with "Not Provided"
data_clean <- data_clean %>%
  mutate(Purchasing.Entity.Type = case_when(
    Purchasing.Entity.Type %in% c("For-Profit Business Corporation", "For Profit Busines Corporation", "For-profit corporation") ~ "For-Profit Organization",
    Purchasing.Entity.Type == "Individual" ~ "Sole Proprietor/Individual",
    is.na(Purchasing.Entity.Type) | Purchasing.Entity.Type == "" ~ "Not Provided", # Include previous cleaning step
    TRUE ~ Purchasing.Entity.Type  # Leave other entries unchanged
  ))

#take changes made further down and put them all here in the intro/setup section
person_lot_count <- data_clean %>%
  group_by(Person.Count) %>%
  distinct(Name.Key) %>%
  summarise(Count_of_People = n_distinct(Name.Key),           # Count distinct people by Name.Key
            Lots_Accounted_For = sum(Person.Count)) %>%  # Calculate total lots
  arrange(Person.Count)

# Create a new variable for the count of lots applied for each Name & Entity Type combo
data_clean <- data_clean %>%
  group_by(Name.Key, Purchasing.Entity.Type) %>%
  mutate(Lot_Application_Count = n()) %>%  # New variable for the count of lots per Person & Entity Type combo
  ungroup()

# Add a column indicating if the entity applied for 1 lot or more than 1 lot based on this new variable
data_clean <- data_clean %>%
  mutate(Lot_Application_Group = case_when(
    Lot_Application_Count == 1 ~ "1 lot",
    Lot_Application_Count > 1 ~ ">1 lot"
  ))

# Convert Date.Submitted to Date format
data_clean <- data_clean %>%
  mutate(Date.Submitted = as.Date(Date.Submitted, format = "%m/%d/%Y"))


# Create a new column for "Sole Proprietor/Individual" or "All Others"
data_clean <- data_clean %>%
  mutate(Entity_Type_Group = case_when(
    Purchasing.Entity.Type == "Sole Proprietor/Individual" ~ "Sole Proprietor/Individual",
    TRUE ~ "All Others"
  ))

# Filter the data for each window and create a categorical variable for each window
data_clean <- data_clean %>%
  mutate(Application_Window = case_when(
    Date.Submitted >= as.Date("2022-11-01") & Date.Submitted <= as.Date("2023-05-31") ~ "Window 1, Nov '22 - May '23",
    Date.Submitted >= as.Date("2023-11-01") & Date.Submitted <= as.Date("2024-01-31") ~ "Window 2, Nov '23 - Jan '24",
    Date.Submitted >= as.Date("2024-03-01") & Date.Submitted <= as.Date("2024-07-31") ~ "Window 3, March '24 - July '24",
    TRUE ~ NA_character_  # Exclude other dates
  ))

#deomgraphic data, really only applies to individuals but filling in for everyone here

data_clean <- data_clean %>%
  mutate(Applicant.Race = ifelse(is.na(Applicant.Race) | Applicant.Race == "", "Not Provided", Applicant.Race), 
         Gender.Identity = ifelse(is.na(Gender.Identity) | Gender.Identity == "", "Not Provided", Gender.Identity), 
         hispanic = ifelse(is.na(hispanic) | hispanic == "", "Not Provided", hispanic), 
         hispanic = ifelse(hispanic == "Not Hispanic or Latino/ Latina/ Latinx", 
                           "Not Hispanic or Latino/Latina/Latinx", hispanic), 
         LGBTQIA = ifelse(is.na(LGBTQIA) | LGBTQIA == "", "Not Provided", LGBTQIA), 
         age = case_when(age == "18-24" ~ "18 - 24", age == "25-34" ~ "25 - 34", age == "35-44" ~ "35 - 44",
          age == "45-54" ~ "45 - 54", TRUE ~ age),  # Leave all other age values unchanged
         age = ifelse(is.na(age) | age == "", "Not Provided", age), 
         income = as.character(income),   # Convert to character
         income = str_trim(income),       # Trim whitespace
         income = ifelse(income %in% c("NA", "") | is.na(income), "Not Provided", income), # Replace "NA" text, blanks, actual NAs
         connection2 = ifelse(connection2 == "10-20 years", "11-20 years", connection2), 
         connection2 = ifelse(is.na(connection2) | connection2 == "", "Not Provided", connection2)
         )

# Extract Property ZIP Code to capture 5 digits before ', USA'
#data_clean$Property_ZipCode <- str_extract(data_clean$address, "\\d{5}(?=, usa)")

# Remove blanks in Application.Type (will clean this later w/ Urban Agriculture lots, etc)
#data_clean <- data_clean
#  mutate(Application.Type = ifelse(is.na(Application.Type) | Application.Type == "", "Not Listed", Application.Type))

data_clean <- data_clean %>%
  mutate(Application.Type = ifelse(is.na(Application.Type) | Application.Type == "", "Not Provided", Application.Type))

#write.csv(data_clean, "CBB_Data_Rebuild2.27.csv", row.names = FALSE)

```

## **Intro**

ChiBlockBuilder (CBB) operated 3 rounds of applications windows for available city lots between Nov '22 and Nov '24.
Applicants included individuals, For-Profits and Non-Profits; types of lot developments included open space, side yards, affordable housing, market rate housing and commercial development.

CBB data exports produced 2,266 lines of applicant info and were provided by City to GCI in August '24.
Once data was cleaned & multi-PIN applications were split up, total records numbered 2,359.

Our first look at the data examines by name of applicant how many lots were applied for.

```{r}
person_lot_count %>%
  kable(col.names = c("Number of Lots Applied For", 
                      "Count of People", 
                      "Total Lots Accounted For"),
        caption = "Count of Lots Applied For by Individual Name")

```

This table groups applicants by the number of lots applied for based on each individual name.
Line one tells us 864 individuals applied for only one lot.
Most individual names are only attached to 1 or a handful of lot applications, although there do appear to be a few "power users" of the application platform.

```{r eval=FALSE}
# Inconsistent data check section - checking for individuals if they have provided different demographic info across multiple applications they've been a part of. 

data_clean_indiv_check <- data_clean %>%
  filter(Purchasing.Entity.Type == "Sole Proprietor/Individual")

# Filter the data to include only individuals with multiple applications
multi_applicants <- data_clean_indiv_check %>%
  group_by(Name.Key) %>%
  filter(n() > 1) %>%
  ungroup() %>%
  select(Record.ID, Name.Key, Applicant.ZIP.Code, connection2, Applicant.Race, hispanic, Gender.Identity, LGBTQIA, age, income, Applicant.Address.1, Applicant.Address.2, Property.Community.Area)

# Define a function to check for inconsistencies in each demographic variable
check_inconsistencies <- function(data, field) {
  # Group by Name.Key and check for cases with both non-missing and missing (NA or "")
  inconsistent <- data %>%
    group_by(Name.Key) %>%
    summarise(inconsistent = any(!is.na(get(field)) & get(field) != "") & any(is.na(get(field)) | get(field) == "")) %>%
    filter(inconsistent) %>%
    pull(Name.Key)  
  return(inconsistent)
}

# List of demographic fields to check
demographic_fields <- c("Applicant.Race", "hispanic", "Gender.Identity", "LGBTQIA", "age", "income")

# Create a list to store the results for each field
inconsistencies <- list()

# Loop through each demographic field and store inconsistent Name.Keys
for (field in demographic_fields) {
  inconsistencies[[field]] <- check_inconsistencies(multi_applicants, field)
}

# Print results
#for (field in names(inconsistencies)) {
#  cat(paste("Inconsistent entries for", field, ":", length(inconsistencies[[field]]), "individuals\n"))
#  print(inconsistencies[[field]])
#}

```

## Purchasing Entity Summary

Each application requests an entity type, responses received are:\
Sole Proprietor/Individual, For-Profit Organization, Non-Profit Organization, Limited Liability Company, Not Provided & Other

Some individual names are attached to applications of multiple entity types.
For example - one of the names applying for 33 lots on the above table submitted 10 applications as a For-Profit Organization, 7 as a Non-Profit Organization, and 16 as a Sole Proprietor/Individual.

```{r, purchasing entity2, warning=FALSE}

ggplot(data_clean, aes(x = fct_infreq(Purchasing.Entity.Type))) +  # Reorder the x-axis
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +  # Add count labels at the top of each bar
  labs(title = "Breakdown of Purchasing Entity Type", 
       x = "Purchasing Entity Type", 
       y = "Count of Applications") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +  # Expand by 10% at the top
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 1: Calculate counts and percentages
entity_counts <- data_clean %>%
  count(Purchasing.Entity.Type) %>%
  mutate(Percent = (n / sum(n)) * 100)  # Calculate percent of total

# Step 2: Create the bar chart using the updated layout
ggplot(entity_counts, aes(x = fct_reorder(Purchasing.Entity.Type, -n), y = Percent)) +  
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Percent)), vjust = -0.5) +  # Percent labels above bars
  geom_text(aes(label = n, y = -5), vjust = 0.5) +  # Raw count below x-axis labels
  labs(title = "Breakdown of Purchasing Entity Type", 
       x = "Purchasing Entity Type", 
       y = "Percent of Applications") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), expand = expansion(mult = c(0.2, 0.5))) +  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1.025),  # Keep labels readable
    axis.title.x = element_text(margin = margin(t = 5)),  # Adjust x-axis title positioning
    plot.margin = margin(t = 10, r = 5, b = 10, l = 50)  # Expand left margin so first label fits
  )


```

The 2 listed as "Other (specify)" provided no additional clarifying information in their application form.

Each application is also tagged with an Application Status of Accepted, In-Progress, Declined or Withdrawn.
I've merged the Declined & Withdrawn status in the following charts and tables.
These values will change over time as more move out of "In-Progress".

```{r, warning=FALSE}
# Group by Application.Status2, calculate counts

#data_clean <- data_clean %>%
#  rename(Application.StatusNull = Application.Status2)

#data_clean <- data_clean %>%
#  rename(Application.Status2 = Application.Status.Output)

status_overall_table <- data_clean %>%
  group_by(Application.Status2) %>%
  summarise(Overall_Count = n())

# Reorder categories in the desired order
status_overall_table$Application.Status2 <- factor(status_overall_table$Application.Status2, 
                                                   levels = c("Accepted", "In-Progress", "Declined & Withdrawn"))

# Create the bar chart
ggplot(status_overall_table, aes(x = Application.Status2, y = Overall_Count)) +
  geom_bar(stat = "identity") +  # Bar chart
  geom_text(aes(label = Overall_Count), vjust = -0.5, size = 5) +  # Add labels above bars
  labs(title = "Application Status Overview",
       x = "Application Status",
       y = "Count") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +  # Expand by 10% at the top
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend since colors represent categories directly


# Create the bar chart
ggplot(status_overall_table, aes(x = Application.Status2, y = Overall_Count, fill = Application.Status2)) +
  geom_bar(stat = "identity") +  # Bar chart
  geom_text(aes(label = Overall_Count), vjust = -0.5, size = 5) +  # Add labels above bars
  labs(title = "Application Status Overview",
       x = "Application Status",
       y = "Count") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +  # Expand by 10% at the top
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend since colors represent categories directly

# Step 1: Calculate total counts and percentages
status_overall_table <- status_overall_table %>%
  mutate(Percent = (Overall_Count / sum(Overall_Count)) * 100)  # Calculate percent of total

# Step 2: Create the bar chart with corrected scale
ggplot(status_overall_table, aes(x = Application.Status2, y = Percent)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = sprintf("%.1f%%", Percent)), vjust = -0.5, size = 5) +  # Show % above bars
  geom_text(aes(label = Overall_Count, y = -5), vjust = .5, size = 4) +  # Bring raw counts back into view
  labs(title = "Application Status Overview",
       x = "Application Status",
       y = "Percent of Applications") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     limits = c(-10, 100),  # Keep scale within 0-100%
                     expand = expansion(mult = c(0, 0.1))) +  # Prevents excess space
  theme_minimal() +
  theme(
    legend.position = "none",  # Remove legend since color is directly shown
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1.025),  # Adjust x-axis labels
    axis.title.x = element_text(margin = margin(t = 5)),  # Adjust x-axis title position
    plot.margin = margin(t = 10, r = 5, b = 10, l = 50)  # Expand left margin to fit first label
  )


```

```{r, warning=FALSE}
# Group by Application.Status2, calculate total
status_overall_table <- status_overall_table %>%
  mutate(percent = (Overall_Count / sum(Overall_Count)) * 100)
  
# Create the bar chart
ggplot(status_overall_table, aes(x = Application.Status2, y = Overall_Count, fill = Application.Status2)) +
  geom_bar(stat = "identity") +  # Bar chart
  geom_text(aes(label = Overall_Count), vjust = -0.5, size = 5) +  # Add labels above bars
  labs(title = "Application Status Overview",
       x = "Application Status",
       y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend since colors represent categories directly


```

```{r, warning=FALSE}
# Group by Purchasing.Entity.Type and Application.Status, and calculate counts and percentages
status_percent_table <- data_clean %>%
  group_by(Purchasing.Entity.Type, Application.Status2) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  ungroup()

# Pivot the table to get Application.Status as columns
status_pivot_table <- status_percent_table %>%
  dplyr::select(Purchasing.Entity.Type, Application.Status2, Percentage) %>%
  pivot_wider(names_from = Application.Status2, values_from = Percentage, values_fill = 0)

# Apply rounding only to percentage columns
status_pivot_table <- status_pivot_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))  # Round all numeric columns to 2 decimals

# Reorder columns: Accepted, In-Progress, Withdrawn, Declined
status_pivot_table <- status_pivot_table %>%
  dplyr::select(Purchasing.Entity.Type, Accepted, `In-Progress`, `Declined & Withdrawn`)

# Display the final table
status_pivot_table %>%
  kable(col.names = c("Entity Type", "Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)"),
        caption = "Percent of Application Status by Purchasing Entity Type")
```

```{r, warning=FALSE}
# Group by Purchasing.Entity.Type and Application.Status, calculate counts
status_count_table <- data_clean %>%
  group_by(Purchasing.Entity.Type, Application.Status2) %>%
  summarise(Count = n()) %>%
  ungroup()

# Pivot the table to get Application.Status as columns (for raw counts)
status_pivot_count_table <- status_count_table %>%
  dplyr::select(Purchasing.Entity.Type, Application.Status2, Count) %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Reorder columns: Accepted, In-Progress, Withdrawn, Declined
status_pivot_count_table <- status_pivot_count_table %>%
  dplyr::select(Purchasing.Entity.Type, Accepted, `In-Progress`, `Declined & Withdrawn`)

# Display the final table with raw counts
status_pivot_count_table %>%
  kable(col.names = c("Entity Type", "Accepted", "In-Progress", "Declined & Withdrawn"),
        caption = "Count of Application Status by Purchasing Entity")
```

```{r, warning=FALSE}
# Group by Purchasing.Entity.Type and Application.Status, and calculate counts
status_percent_table <- data_clean %>%
  group_by(Purchasing.Entity.Type, Application.Status2) %>%
  summarise(Count = n()) %>%
  ungroup()

# Pivot the table to get Application.Status as columns and add a row at the end for totals
status_pivot_table <- status_percent_table %>%
  dplyr::select(Purchasing.Entity.Type, Application.Status2, Count) %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0) %>%
  adorn_totals() 

#Create a row that sums together the application counts for all status for each row (entity type)
status_pivot_table <- status_pivot_table %>%
  mutate(row_sum = rowSums(status_pivot_table[,2:4]), 
         percent_accepted = (Accepted / row_sum) * 100,
         percent_declinedwithdrawn = (`Declined & Withdrawn` / row_sum) * 100,
         percent_inprog = (`In-Progress` / row_sum) * 100)

# Apply rounding only to percentage columns
status_pivot_table <- status_pivot_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))  # Round all numeric columns to 2 decimals

status_pivot_table <- status_pivot_table %>%
  select(Purchasing.Entity.Type, percent_accepted, percent_declinedwithdrawn, percent_inprog)

# Reorder columns: Accepted, In-Progress, Withdrawn, Declined
status_pivot_table <- status_pivot_table %>%
  dplyr::select(Purchasing.Entity.Type, percent_accepted, percent_inprog, percent_declinedwithdrawn)

# Display the final table
status_pivot_table %>%
  kable(col.names = c("Entity Type", "Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)"),
        caption = "Percent of Application Status by Purchasing Entity Type") 
```

Calculating the percent within each group that are Approved vs In-Progress vs Declined & Withdrawn tells us which types of applications are more likely to receive approval.
Notable is that the LLCs have a considerably higher approval rate compared to the other entity types (more on this further down).
While the other entity types have a slight variance in their Accepted %'s, the "Declined & Withdrawn" percentages are similar.LLCs also have a notably higher acceptance rate than the acceptance rate for all applications.
All other entity types are within two or three percentage points of the overall acceptance rate.

```{r}
# Create a stacked bar chart for Purchasing.Entity.Type stacked by Application.Status
ggplot(data_clean, aes(x = fct_infreq(Purchasing.Entity.Type), fill = Application.Status2)) +
  geom_bar() +
  labs(title = "Purchasing Entity Type Breakdown Volume by Application Status", 
       x = "Purchasing Entity Type", 
       y = "Count of Applications",
       fill = "Application Status") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +  # Expand by 10% at the top
  scale_fill_manual(values = c("Accepted" = "lightgreen", "Declined & Withdrawn" = "pink", "In-Progress" = "skyblue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Visualizing the status by entity type doesn't uncover additional irregularities in the data.

We wanted to explore if there is a difference in acceptance rates between entities applying for one lot vs those with more than one application.
Individual/Sole Proprietors were filtered out of this set.
The first table showing counts is less important than the second showing %'s.

```{r}
# Filter out Sole Proprietor/Individual from the analysis as requested
data_clean_business <- data_clean %>%
  filter(Purchasing.Entity.Type != "Sole Proprietor/Individual")


# Step 3: Group by Purchasing.Entity.Type, Lot_Application_Group, and Application.Status, and calculate counts
status_count_table <- data_clean_business %>%
  group_by(Purchasing.Entity.Type, Lot_Application_Group, Application.Status2) %>%
  summarise(Count = n()) %>%
  ungroup()

# Pivot the table to get Application.Status as columns (for raw counts)
status_pivot_count_table <- status_count_table %>%
  dplyr::select(Purchasing.Entity.Type, Lot_Application_Group, Application.Status2, Count) %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Reorder columns: Accepted, In-Progress, Withdrawn, Declined
status_pivot_count_table <- status_pivot_count_table %>%
  dplyr::select(Purchasing.Entity.Type, Lot_Application_Group, Accepted, `In-Progress`, `Declined & Withdrawn`)

# Display the final table with raw counts
status_pivot_count_table %>%
  kable(col.names = c("Entity Type", "Lot Application Group", "Accepted", "In-Progress", "Declined & Withdrawn"),
        caption = "Count of Application Status within Purchasing Entity Groups (1, 1+)")

# Step 4: Percentage Breakdown Table
status_percent_table <- data_clean_business %>%
  group_by(Purchasing.Entity.Type, Lot_Application_Group, Application.Status2) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  ungroup()

# Pivot the table to get Application.Status as columns (for percentages)
status_pivot_table <- status_percent_table %>%
  dplyr::select(Purchasing.Entity.Type, Lot_Application_Group, Application.Status2, Percentage) %>%
  pivot_wider(names_from = Application.Status2, values_from = Percentage, values_fill = 0)

# Apply rounding only to percentage columns
status_pivot_table <- status_pivot_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))  # Round all numeric columns to 2 decimals

# Reorder columns: Accepted, In-Progress, Withdrawn, Declined
status_pivot_table <- status_pivot_table %>%
  dplyr::select(Purchasing.Entity.Type, Lot_Application_Group, Accepted, `In-Progress`, `Declined & Withdrawn`)

# Display the final table with percentages
status_pivot_table %>%
  kable(col.names = c("Entity Type", "Lot Application Group", "Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)"),
        caption = "Percent of Application Status within Purchasing Entity Group (1, >1)")

```

The LLCs behaving differently from the other types is also clear on this table, just in terms of acceptance rate.
None of the other entity types show a stark difference in acceptance by organizations that are able to apply for multiple lots, implying there's no inherent advantage for the larger groups in the program.
In each non-LLC entity type the acceptance rate of those with more than 1 lot application is actually lower than those with only 1 lot.
There are still a number of "In-Progress" applications for all types here, so it's possible this observation could change with updated data.

## Property Application Type

```{r, property type}
# Step 2: Define the correct order for Application.Type
application_type_order <- c("Open Space", "Side Yard", "Market Rate Housing", "Affordable Housing", "Commercial Development", "Urban Agriculture", "Other")

data_clean <- data_clean %>%
  mutate(Application.Type = factor(Application.Type, levels = application_type_order))

# Step 3: Basic Bar Chart for Application.Type
ggplot(data_clean, aes(x = Application.Type)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +  # Add count labels at the top of each bar
  labs(title = "Count of Applications by Type", 
       x = "Application Type", 
       y = "Count of Applications") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +  # Expand by 20% at the top
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 1: Calculate total counts and percentages
application_type_counts <- data_clean %>%
  count(Application.Type) %>%
  mutate(Percent = (n / sum(n)) * 100)  # Calculate percent of total

# Step 2: Create the bar chart with adjusted layout
ggplot(application_type_counts, aes(x = fct_reorder(Application.Type, -n), y = Percent)) +  
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Percent)), vjust = -0.5) +  # Show % above bars
  geom_text(aes(label = n, y = 0), vjust = 1.5) +  # Keep raw count above the x-axis line
  labs(title = "Count of Applications by Type", 
       x = "Application Type", 
       y = "Percent of Applications") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     expand = expansion(mult = c(0.2, 0.5), add = c(0, 0))) +  # Prevents negative y-axis
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1.025),  # Adjust x-axis labels
    axis.title.x = element_text(margin = margin(t = 5)),  # Adjust x-axis title position
    plot.margin = margin(t = 10, r = 5, b = 10, l = 50)  # Expand left margin to fit first label
  )

```

The Application Type charts reveal there are still a large number of lots applied for with no specific "use type" tagged in the system (502).
Revisiting the raw data uncovers that a small portion (40) of these were attached to two "Urban Agriculture" application forms included in Windows 2 & 3.
The remaining 462 records may have a use type present in the CBB in a record that was not provided to GCI.

```{r, property type3}
# Step 4: Stacked Bar Chart for Application Type by Application Status
ggplot(data_clean, aes(x = Application.Type, fill = Application.Status2)) +
  geom_bar(position = "stack") +
  labs(title = "Application Type Stacked by Application Status", 
       x = "Application Type", 
       y = "Count of Applications", 
       fill = "Application Status") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +  # Expand by 20% at the top
  scale_fill_manual(values = c("Accepted" = "lightgreen", "Declined & Withdrawn" = "pink", "In-Progress" = "skyblue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Viewing the Application Types by Status reveals Side Yards as having the highest concentration of Accepteds.
Both Market Rate Housing & "Not Listed" have high volumes of In-Progress applications (in blue).
The "Not Listed" group makes up most of the two later rounds of applications so the higher presence of In-Progress makes sense for more recent applications.
Gaining additional data that clarifies the intention of the applicants for those lots will be useful.

```{r warning=FALSE}
# Group by Application.Type and Application.Status, calculate counts
status_type_count_table <- data_clean %>%
  group_by(Application.Type, Application.Status2) %>%
  summarise(Count = n()) %>%
  ungroup()

# Pivot the table to get Application.Status as columns (for raw counts)
status_type_pivot_count_table <- status_type_count_table %>%
  dplyr::select(Application.Type, Application.Status2, Count) %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Reorder columns: Accepted, In-Progress, Withdrawn, Declined
status_type_pivot_count_table <- status_type_pivot_count_table %>%
  dplyr::select(Application.Type, Accepted, `In-Progress`, `Declined & Withdrawn`)

# Display the final table with raw counts
status_type_pivot_count_table %>%
  kable(col.names = c("Application Type", "Accepted", "In-Progress", "Declined & Withdrawn"),
        caption = "Count of Application Status by Lot Application Type")
```

In the data provided, only 33 lot sales involving a built structure have been approved so far (21 market rate housing, 11 affordable housing & 1 commercial development).
It's possible some of the 45 approved "not listed" developments are types other than side yard or open space and they will increase that count.

```{r warning=FALSE}
# Group by Application.Type and Application.Status, and calculate counts and percentages
status_type_percent_table <- data_clean %>%
  group_by(Application.Type, Application.Status2) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  ungroup()

# Pivot the table to get Application.Status as columns
status_type_pivot_table <- status_type_percent_table %>%
  dplyr::select(Application.Type, Application.Status2, Percentage) %>%
  pivot_wider(names_from = Application.Status2, values_from = Percentage, values_fill = 0)

# Apply rounding only to percentage columns
status_type_pivot_table <- status_type_pivot_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))  # Round all numeric columns to 2 decimals

# Reorder columns: Accepted, In-Progress, Withdrawn, Declined
status_type_pivot_table <- status_type_pivot_table %>%
  dplyr::select(Application.Type, Accepted, `In-Progress`, `Declined & Withdrawn`)

# Display the final table
status_type_pivot_table %>%
  kable(col.names = c("Application Type", "Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)"),
        caption = "Percent of Application Status by Lot Application Type")
```

```{r, warning=FALSE}
# Group by Application.Type and Application.Status, and calculate counts
status_percent_table <- data_clean %>%
  group_by(Application.Type, Application.Status2) %>%
  summarise(Count = n()) %>%
  ungroup()

# Pivot the table to get Application.Status as columns and add a row at the end for totals
status_pivot_table <- status_percent_table %>%
  dplyr::select(Application.Type, Application.Status2, Count) %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0) %>%
  adorn_totals() 

#Create a row that sums together the application counts for all status for each row (entity type)
status_pivot_table <- status_pivot_table %>%
  mutate(row_sum = rowSums(status_pivot_table[,2:4]), 
         percent_accepted = (Accepted / row_sum) * 100,
         percent_declinedwithdrawn = (`Declined & Withdrawn` / row_sum) * 100,
         percent_inprog = (`In-Progress` / row_sum) * 100)

# Apply rounding only to percentage columns
status_pivot_table <- status_pivot_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))  # Round all numeric columns to 2 decimals

status_pivot_table <- status_pivot_table %>%
  select(Application.Type, percent_accepted, percent_declinedwithdrawn, percent_inprog)


# Reorder columns: Accepted, In-Progress, Withdrawn, Declined
status_pivot_table <- status_pivot_table %>%
  dplyr::select(Application.Type, percent_accepted, percent_inprog, percent_declinedwithdrawn)

# Display the final table
status_pivot_table %>%
  kable(col.names = c("Entity Type", "Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)"),
        caption = "Percent of Application Status by Lot Application Type") 
```

```{r}
# Step 1: Create a summary table by grouping on Application.Type and Purchasing.Entity.Type
application_type_summary <- data_clean %>%
  group_by(Application.Type, Purchasing.Entity.Type) %>%
  summarise(Count = n(), .groups = "drop")

# Step 2: Pivot the table to create a crosstable
application_type_crosstable <- application_type_summary %>%
  pivot_wider(names_from = Purchasing.Entity.Type, values_from = Count, values_fill = 0)

# Step 3: Add percentage columns, rounding to 1 decimal place
application_type_percent <- application_type_crosstable %>%
  mutate(across(-Application.Type, ~ round((.x / sum(.x)) * 100, 1), .names = "{.col} (%)"))

# Step 4: Convert all percentage columns to character with "%" suffix
application_type_percent <- application_type_percent %>%
  mutate(across(ends_with("(%)"), ~ paste0(.x, "%"), .names = "{.col}"))

# Step 5: Manually create the Totals Row
totals_row <- application_type_crosstable %>%
  summarise(across(-Application.Type, sum)) %>%
  mutate(Application.Type = "Total")

# Step 6: Ensure percentage totals are formatted correctly
for (entity in colnames(application_type_crosstable)[-1]) {  # Skip "Application.Type"
  percent_col <- paste0(entity, " (%)")
  if (percent_col %in% colnames(application_type_percent)) {
    totals_row[[percent_col]] <- "100.0%"  # Assign consistent character format
  }
}

# Step 7: Convert all percentage columns in totals_row to character type before binding
totals_row <- totals_row %>%
  mutate(across(ends_with("(%)"), as.character))

# Step 8: Bind the Totals Row to the dataset
application_type_final <- bind_rows(application_type_percent, totals_row)

# Step 9: Ensure correct column order (Count next to Percent)
ordered_columns <- c("Application.Type", 
                     rbind(colnames(application_type_crosstable)[-1], 
                           paste0(colnames(application_type_crosstable)[-1], " (%)"))) %>% as.vector()

application_type_final <- application_type_final %>%
  dplyr::select(all_of(ordered_columns))  # Ensure correct column order

# Define entity types in the correct order
entity_types <- c("For-Profit Organization", "Non-Profit Organization", "Sole Proprietor/Individual", 
                  "Limited Liability Company (LLC)", "Not Provided")


# Format and Display the Table
application_type_final %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c"), col.names = NULL) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  add_header_above(c("Application Type" = 1, rep(c("Count", "Percent"), length(entity_types)))) %>%
  add_header_above(c(" " = 1, "For-Profit Organization" = 2, "Non-Profit Organization" = 2, 
                     "Sole Proprietor/Individual" = 2, "Limited Liability Company (LLC)" = 2, "Not Provided" = 2)) %>%
  column_spec(2:11, width = "7em", extra_css = "text-align: right;") %>%
  row_spec(nrow(application_type_final), bold = TRUE)  # Bold the totals row

```

## Date Submitted/Application Window

There are three rounds of ChiBlockBuilder applications represented in this dataset.
The first was officially open from 11/17/22 - 2/3/23, although the dataset provided shows "date submitted" values stretching through the end of May '23.
The "date submitted" field also shows an overwhelmingly large number of the applications from this round on the last day of the window (2/3/23).
We believe this field is being utilized internally to track updates to applications and that is not the actual date submitted for most of those.

```{r, phase timelines}

# Step 4: Filter out rows that are not part of any window
data_clean_window <- data_clean %>%
  filter(!is.na(Application_Window))

# Step 5: Group by Date.Submitted, Application_Window, and Entity_Type_Group to count the number of lots per day in each window
daily_lots_by_window <- data_clean_window %>%
  group_by(Date.Submitted, Application_Window, Entity_Type_Group) %>%
  summarise(Lots_Per_Day = n()) %>%
  ungroup()

# Step 6: Create separate bar charts for each window
window_1 <- ggplot(daily_lots_by_window %>% filter(Application_Window == "Window 1, Nov '22 - May '23"),
                   aes(x = Date.Submitted, y = Lots_Per_Day, fill = Entity_Type_Group)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Sole Proprietor/Individual" = "#1f77b4", "All Others" = "#ff7f0e")) +  # Custom colors
  labs(title = "Window 1: Nov '22 - May '23", x = "Date Submitted", y = "Number of Lots Submitted", fill = "Entity Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

window_2 <- ggplot(daily_lots_by_window %>% filter(Application_Window == "Window 2, Nov '23 - Jan '24"),
                   aes(x = Date.Submitted, y = Lots_Per_Day, fill = Entity_Type_Group)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Sole Proprietor/Individual" = "#1f77b4", "All Others" = "#ff7f0e")) +
  labs(title = "Window 2: Nov '23 - Jan '24", x = "Date Submitted", y = "Number of Lots Submitted", fill = "Entity Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

window_3 <- ggplot(daily_lots_by_window %>% filter(Application_Window == "Window 3, March '24 - July '24"),
                   aes(x = Date.Submitted, y = Lots_Per_Day, fill = Entity_Type_Group)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Sole Proprietor/Individual" = "#1f77b4", "All Others" = "#ff7f0e")) +
  labs(title = "Window 3: March '24 - July '24", x = "Date Submitted", y = "Number of Lots Submitted", fill = "Entity Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Window 1: Nov '22 - May '23
entity_table_window1 <- data_clean_window %>%
  filter(Application_Window == "Window 1, Nov '22 - May '23") %>%
  group_by(Purchasing.Entity.Type) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

application_table_window1 <- data_clean_window %>%
  filter(Application_Window == "Window 1, Nov '22 - May '23") %>%
  group_by(Application.Type) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

# Display Window 1 chart and tables
print(window_1)
```

1,797 lots were applied for in Phase 1, accounting for 76.8% of the lots between the 3 phases.

```{r, phase timelines2}
entity_table_window1 %>%
  kable(col.names = c("Purchasing Entity Type", "Count"), caption = "Entity Type Summary for Window 1")
application_table_window1 %>%
  kable(col.names = c("Application Lot Type", "Count"), caption = "Lot Type Summary for Window 1")

```

All of the applications from Phase 1 have an Application Type identified.
While 167 of the Purchasing Entities from Phase 1 are "Not Provided", the majority of the entities in this round are identified.

116 of the total 121 LLCs are represented in Phase 1.
Only one of those LLCs from Phase 1 is not an application for a Side Yard.

```{r, window 2}
# Window 2: Nov '23 - Jan '24
entity_table_window2 <- data_clean_window %>%
  filter(Application_Window == "Window 2, Nov '23 - Jan '24") %>%
  group_by(Purchasing.Entity.Type) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

application_table_window2 <- data_clean_window %>%
  filter(Application_Window == "Window 2, Nov '23 - Jan '24") %>%
  group_by(Application.Type) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

# Display Window 2 chart and tables
print(window_2)
entity_table_window2 %>%
  kable(col.names = c("Purchasing Entity Type", "Count"), caption = "Entity Type Summary for Window 2")
application_table_window2 %>%
  kable(col.names = c("Application Lot Type", "Count"), caption = "Lot Type Summary for Window 2")
```

114 lots were applied for in Phase 2 (4.8%), and the majority of them do not have a lot type identified.

```{r, window 3}
# Window 3: March '24 - July '24
entity_table_window3 <- data_clean_window %>%
  filter(Application_Window == "Window 3, March '24 - July '24") %>%
  group_by(Purchasing.Entity.Type) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

application_table_window3 <- data_clean_window %>%
  filter(Application_Window == "Window 3, March '24 - July '24") %>%
  group_by(Application.Type) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

# Display Window 3 chart and tables
print(window_3)
entity_table_window3 %>%
  kable(col.names = c("Purchasing Entity Type", "Count"), caption = "Entity Type Summary for Window 3")
application_table_window3 %>%
  kable(col.names = c("Application Lot Type", "Count"), caption = "Lot Type Summary for Window 3")

```

448 lots were applied for in Phase 3 (19%), and again the majority of them do not have a lot type identified.

# Demographic Data of Applicants

## Unique Individuals

```{r unique names count, eval=FALSE}
unique_names_count <- n_distinct(data_clean$Name.Key)
unique_names_count  # This will print the total number of unique names

# Filter to Sole Proprietor/Individual and calculate the number of unique names
unique_solep_count <- data_clean %>%
  filter(Purchasing.Entity.Type == "Sole Proprietor/Individual") %>%
  summarise(unique_names = n_distinct(Name.Key))

unique_solep_count  # This will print the number of unique names

```

There are 1,261 unique names making up the 2,359 applications, although many of those are attached to organizations.
Filtering to only the sole proprietors/individuals reduces the number of unique individuals to 860.

The following demographic charts cover those 860 applicants who applied as "Individual/Sole Proprietor", and they account for 1,339 lots.
598 individuals consistently did not provide any demographic responses in this section on their applications.

The demographic data provided is not always consistent, some applicants with multiple entries may have provided conflicting information on individual applications.

First, a check to see if the 598 individuals not providing any demographic responses are concentrated within specific application windows.

```{r}

data_clean_indiv <- data_clean %>%
  filter(Purchasing.Entity.Type == "Sole Proprietor/Individual")

#test round distribution of Not Provided individuals

data_clean_notprovided <- data_clean_indiv %>%
  filter(Gender.Identity == "Not Provided")

data_clean_notprovided_summary <- data_clean_notprovided %>%
  group_by(Application_Window) %>%
  summarise(Count = n())  # Count entries in each window

ggplot(data_clean_notprovided_summary, aes(x = Application_Window, y = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5) +  # Add count labels above the bars
  labs(title = "Not Provided distribution across 3 Application Windows", 
       x = "Application Window", 
       y = "Count of Responses") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +  # Expand by 10% at the top
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Explore Acceptance Rate of Demographic "Not Provided" vs those that did provide for discrepancies.

```{r}
# Step 1: Create an identifier for whether demographic info was provided
data_clean <- data_clean %>%
  mutate(Demos.Provided = if_else(Gender.Identity == "Not Provided", "Not Provided", "Provided"))

# Step 2: Filter to Sole Proprietor/Individual applications only
sole_prop_data <- data_clean %>%
  filter(Purchasing.Entity.Type == "Sole Proprietor/Individual")

# Step 3: Group by Demos.Provided and Application.Status2, then calculate counts
status_demo_table <- sole_prop_data %>%
  group_by(Demos.Provided, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 4: Pivot wider so each Application.Status2 is a separate column
status_demo_pivot <- status_demo_table %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0) %>%
  adorn_totals(name = "Total", where = "row") # Add total row for context

# Step 5: Calculate row-wise percentages
status_demo_pivot <- status_demo_pivot %>%
  mutate(
    row_sum = rowSums(select(., -Demos.Provided)),  # Total applications per row
    `Accepted (%)` = (Accepted / row_sum) * 100,
    `In-Progress (%)` = (`In-Progress` / row_sum) * 100,
    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / row_sum) * 100,
    `Total (%)` = 100  # Since total column should always be 100%
  ) %>%
  select(Demos.Provided, 
         Accepted, `Accepted (%)`, 
         `In-Progress`, `In-Progress (%)`, 
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`, 
         row_sum, `Total (%)`)  # Keep total counts and percentages at the end

# Step 6: Round values for presentation and ensure percentage formatting
status_demo_pivot <- status_demo_pivot %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%  # Round to 1 decimal
  mutate(across(ends_with("(%)"), ~ paste0(.x, "%")))  # Append "%" to percentage columns

# Step 7: Display as a table
status_demo_pivot %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), 
    col.names = c("Demographic Info Provided", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Total", "Percent"),
    caption = "Application Status for Sole Proprietor/Individual by Demographic Information Provided"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  add_header_above(c(" " = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%
  column_spec(2:9, width = "7em", extra_css = "text-align: right;") %>%
  row_spec(nrow(status_demo_pivot), bold = TRUE)  # Bold the totals row

```

All Sole Proprietor/Individuals had demographic responses.
"Not Provided" exploration no longer valid.

```{r}
# Step 1: Group by Demos.Provided and Application.Type to count applications
app_type_demo_table <- sole_prop_data %>%
  group_by(Demos.Provided, Application.Type) %>%
  summarise(Total_Applications = n(), .groups = "drop")

# Step 2: Calculate percentages within each Demos.Provided group
app_type_demo_table <- app_type_demo_table %>%
  group_by(Demos.Provided) %>%
  mutate(Percent = (Total_Applications / sum(Total_Applications)) * 100) %>%
  ungroup()

# Step 3: Pivot wider so that Provided & Not Provided are columns
app_type_pivot <- app_type_demo_table %>%
  pivot_wider(names_from = Demos.Provided, values_from = c(Total_Applications, Percent), values_fill = 0)

# Step 4: Round percentage values
app_type_pivot <- app_type_pivot %>%
  mutate(across(starts_with("Percent_"), ~ round(.x, 2)))

# Step 5: Display as a table -- TABLE NO LONGER NECESSARY
#app_type_pivot %>%
#  kable(col.names = c("Application Type", 
#                      "Total Not Provided", "Total Provided", 
#                      "% Not Provided", "% Provided"),
#        caption = "Application Type Breakdown by Demographic Info Provided")
```

```{r}
# Step 1: Count applications by Application.Type, Demos.Provided, and Application.Status2
app_status_table <- sole_prop_data %>%
  group_by(Application.Type, Demos.Provided, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 2: Pivot the table to get counts of Accepted, In-Progress, and Declined & Withdrawn
app_status_pivot <- app_status_table %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Step 3: Calculate acceptance rate
app_status_pivot <- app_status_pivot %>%
  mutate(Total = rowSums(select(., -Application.Type, -Demos.Provided)),
         Acceptance_Rate = (Accepted / Total) * 100) %>%
  select(Application.Type, Demos.Provided, Acceptance_Rate)

# Step 4: Pivot wider so that Demos.Provided groups are columns
acceptance_pivot <- app_status_pivot %>%
  pivot_wider(names_from = Demos.Provided, values_from = Acceptance_Rate, values_fill = NA)

# Step 5: Round percentages
acceptance_pivot <- acceptance_pivot %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Step 6: Display table with correct column labels -- NO LONGER NEEDED
#acceptance_pivot %>%
#  kable(col.names = c("Application Type", "Accepted % (Not Provided)", "Accepted % (Provided)"),
#        caption = "Acceptance Rate by Application Type and Demographic Info Provided")

```

### Race

```{r}

unique_name_race_count <- data_clean_indiv %>%
  distinct(Name.Key, Applicant.Race) %>%
  count()
#unique_name_race_count
```

Of the 809 individuals there are 835 name:race combinations.
Some of the duplication comes from individuals who selected a race on one application and left the field blank on another.
Some applicants identifying with multiple races selected each response and the field may have represented those selections in the order they were entered.
For example, these 3 entries are the same individual:

Black or African American Native American or Alaskan Native Asian or Pacific Islander

Black or African American Native American or Alaskan Native Asian or Pacific Islander

Black or African American Asian or Pacific Islander Native American or Alaskan Native

We have cleaned the data to account for those irregularities within individuals when the reponses are the same despite having a different order.

```{r}
# Define possible race responses
race_responses <- c("Black or African American", 
                    "Native American or Alaskan Native", 
                    "Asian or Pacific Islander", 
                    "White or Caucasian", 
                    "I prefer not to answer", 
                    "I identify as a race that is not listed here")

# Initialize an empty data frame to store the counts
race_counts <- data.frame(Race = character(), Count = integer(), stringsAsFactors = FALSE)


#filter out "Not Provided"
data_clean_indiv <- data_clean_indiv %>%
  filter(Applicant.Race != "Not Provided")

# Loop over each race response and count occurrences using str_detect
for (race in race_responses) {
  count <- data_clean_indiv %>%
    filter(str_detect(Applicant.Race, race)) %>%
    summarise(Count = n()) %>%
    pull(Count)
  
  race_counts <- rbind(race_counts, data.frame(Race = race, Count = count))
}

# Generate a bar chart to visualize race counts
ggplot(race_counts, aes(x = reorder(Race, -Count), y = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5) +  # Add count labels above the bars
  labs(title = "Race Distribution of Sole Proprietor/Individual Applicants", 
       x = "Race", 
       y = "Count of Responses") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +  # Expand by 10% at the top
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

As individuals can identify as more than one race, the total of the columns above (1,362) exceeds the number of applications filed by sole proprietors/individuals (1,339).

Race Acceptance Rate

```{r}
# Initialize an empty data frame to store results
race_acceptance <- data.frame(Race = character(), Acceptance_Percent = numeric(), stringsAsFactors = FALSE)

# Loop over each race response and calculate acceptance percentage
for (race in race_responses) {
  
  # Filter for total applications matching this race
  total_apps <- data_clean_indiv %>%
    filter(str_detect(Applicant.Race, race)) %>%
    summarise(Count = n()) %>%
    pull(Count)
  
  # Filter for accepted applications matching this race
  accepted_apps <- data_clean_indiv %>%
    filter(str_detect(Applicant.Race, race) & Application.Status2 == "Accepted") %>%
    summarise(Count = n()) %>%
    pull(Count)
  
  # Calculate acceptance rate
  acceptance_rate <- ifelse(total_apps > 0, (accepted_apps / total_apps) * 100, NA)
  
  # Append to results data frame
  race_acceptance <- rbind(race_acceptance, data.frame(Race = race, Acceptance_Percent = round(acceptance_rate, 2)))
}

# Display the results as a table
race_acceptance %>%
  kable(col.names = c("Race", "Acceptance Rate (%)"),
        caption = "Acceptance Rate by Race")

```

### Gender

```{r}
# Define possible gender responses
gender_responses <- c("Female", 
                      "Male", 
                      "Gender Non-Conforming/Non-Binary/Gender Queer", 
                      "Transgender", 
                      "I identify as a gender that is not listed here", 
                      "I prefer not to answer")

# Initialize an empty data frame to store the counts for gender
gender_counts <- data.frame(Gender = character(), Count = integer(), stringsAsFactors = FALSE)

#filter out "Not Provided"
data_clean_indiv <- data_clean_indiv %>%
  filter(Applicant.Race != "Not Provided")

# Loop over each gender response and count occurrences using str_detect
for (gender in gender_responses) {
  count <- data_clean_indiv %>%
    filter(str_detect(Gender.Identity, gender)) %>%
    summarise(Count = n()) %>%
    pull(Count)
  
  gender_counts <- rbind(gender_counts, data.frame(Gender = gender, Count = count))
}

# Generate a bar chart to visualize gender counts
ggplot(gender_counts, aes(x = reorder(Gender, -Count), y = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5) +  # Add count labels above the bars
  labs(title = "Gender Distribution of Sole Proprietor/Individual Applicants", 
       x = "Gender", 
       y = "Count of Responses") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +  # Expand by 10% at the top
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 1: Calculate total counts and percentages
gender_counts <- gender_counts %>%
  mutate(Percent = (Count / sum(Count)) * 100)  # Calculate percent of total

# Step 2: Create the bar chart with improved layout
ggplot(gender_counts, aes(x = fct_reorder(Gender, -Count), y = Percent)) +  
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Percent)), vjust = -0.5) +  # Show % above bars
  geom_text(aes(label = Count, y = 0), vjust = 1.2) +  # Raw count below x-axis labels
  labs(title = "Gender Distribution of Sole Proprietor/Individual Applicants", 
       x = "Gender", 
       y = "Percent of Responses") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     expand = expansion(mult = c(0.2, 0.5), add = c(0, 0))) +  # Prevents negative y-axis
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1.015),  # Lower x-axis labels
    axis.title.x = element_text(margin = margin(t = 5)),  # Adjust x-axis title position
    plot.margin = margin(t = 10, r = 5, b = 10, l = 55)  # Slightly increase left margin
  )


```

Gender Acceptance Rate

```{r}
# Step 1: Assign a single gender category per applicant based on hierarchy
data_clean_indiv <- data_clean_indiv %>%
  mutate(Gender_Simplified = case_when(
    str_detect(Gender.Identity, "Transgender") ~ "Transgender",
    str_detect(Gender.Identity, "Male") ~ "Male",
    str_detect(Gender.Identity, "Female") ~ "Female",
    str_detect(Gender.Identity, "Gender Non-Conforming/Non-Binary/Gender Queer") ~ "Gender Non-Conforming/Non-Binary/Gender Queer",
    str_detect(Gender.Identity, "I identify as a gender that is not listed here") ~ "I identify as a gender that is not listed here",
    str_detect(Gender.Identity, "I prefer not to answer") ~ "I prefer not to answer",
    TRUE ~ "Not Provided"
  ))

# Step 2: Create a summary table by grouping on Gender_Simplified and Application.Status2
gender_status_summary <- data_clean_indiv %>%
  group_by(Gender_Simplified, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 3: Pivot the table to create a crosstable
gender_status_pivot <- gender_status_summary %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Step 4: Calculate total applications per gender category
gender_status_pivot <- gender_status_pivot %>%
  mutate(Total = rowSums(select(., -Gender_Simplified)))

# Step 5: Calculate percentages for each status
gender_status_percent <- gender_status_pivot %>%
  mutate(
    `Accepted (%)` = (Accepted / Total) * 100,
    `In-Progress (%)` = (`In-Progress` / Total) * 100,
    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / Total) * 100,
    `Total (%)` = 100  # Since total column should always be 100%
  ) 

# Step 6: Ensure rounding and percentage formatting
gender_status_percent <- gender_status_percent %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%  # Round to 1 decimal
  mutate(across(ends_with("(%)"), ~ paste0(.x, "%")))  # Append "%" to percentage columns

# Step 7: Manually create the Totals Row
totals_row <- gender_status_pivot %>%
  summarise(across(-Gender_Simplified, sum)) %>%
  mutate(Gender_Simplified = "Total")

# Step 7.1: Append "100.0%" manually to all percentage columns
for (col in c("Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)", "Total (%)")) {
  totals_row[[col]] <- "100.0%"  # Explicitly set to "100.0%" for all status columns
}

# Step 8: Append the Totals Row to the dataset
gender_status_final <- bind_rows(gender_status_percent, totals_row)

# Step 9: Define correct ordering for gender categories
gender_order <- c(
  "Male", 
  "Female", 
  "I prefer not to answer", 
  "Gender Non-Conforming/Non-Binary/Gender Queer", 
  "I identify as a gender that is not listed here", 
  "Transgender",
  "Total"
)

# Ensure proper order
gender_status_final <- gender_status_final %>%
  mutate(Gender_Simplified = factor(Gender_Simplified, levels = gender_order)) %>%
  arrange(Gender_Simplified)

# Step 10: Reorder columns to interleave count and percentage
gender_status_final <- gender_status_final %>%
  select(Gender_Simplified, 
         Accepted, `Accepted (%)`, 
         `In-Progress`, `In-Progress (%)`, 
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`, 
         Total, `Total (%)`)  # Ensure alternating columns

# Step 11: Format and Display the Table
gender_status_final %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), 
    col.names = c("Gender", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Total", "Percent"),
    caption = "Application Status by Gender"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = FALSE, 
                position = "left", 
                font_size = 12,  # Ensures text fits correctly
                latex_options = "scale_down") %>%
  add_header_above(c(" " = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%
  column_spec(1, width = "30em") %>%  # Expands first column width
  column_spec(2:9, width = "7em") %>%  # Ensures readability without forced equal column widths
  row_spec(nrow(gender_status_final), bold = TRUE)  # Bold the totals row

```

### Hispanic

```{r}
# Count the occurrences of each response
hispanic_counts <- data_clean_indiv %>%
  filter(hispanic %in% c("Hispanic or Latino/Latina/Latinx", 
                         "I prefer not to answer", 
                         "Not Hispanic or Latino/Latina/Latinx")) %>%
  group_by(hispanic) %>%
  summarise(Count = n())

#filter out "Not Provided"
data_clean_indiv <- data_clean_indiv %>%
  filter(Applicant.Race != "Not Provided")

# Calculate the percentage of Hispanic (excluding "I prefer not to answer" and "Not Provided")
hispanic_percentage <- data_clean_indiv %>%
  filter(hispanic %in% c("Hispanic or Latino/Latina/Latinx", "Not Hispanic or Latino/Latina/Latinx")) %>%
  summarise(
    Hispanic_Count = sum(hispanic == "Hispanic or Latino/Latina/Latinx"),
    Total_Provided = n(),
    Percent_Hispanic = (Hispanic_Count / Total_Provided) * 100
  ) %>%
  pull(Percent_Hispanic)

# Add the percentage Hispanic row to the dataset
hispanic_counts <- rbind(hispanic_counts, data.frame(hispanic = "% Hispanic (of those who provided response)", Count = round(hispanic_percentage, 2)))

# Reorder the rows 
hispanic_counts <- hispanic_counts %>%
  arrange(factor(hispanic, levels = c("Not Provided", "I prefer not to answer", "Not Hispanic or Latino/Latina/Latinx", "Hispanic or Latino/Latina/Latinx", "% Hispanic (of those who provided response)")))

hispanic_counts <- hispanic_counts %>%
  mutate(Count = ifelse(hispanic == "% Hispanic (of those who provided response)", 
                        format(Count, nsmall = 2), 
                        as.character(Count)))  # Keep other counts as whole numbers

# Use kable to display the hispanic_counts table in a more presentable format
hispanic_counts %>%
  kable(
    caption = "Hispanic Counts and Percentage Breakdown",
    col.names = c("Hispanic Status", "Count"),
    align = c("l", "r"),
    format = "markdown"  # Or "html" for HTML output
  )

```

Hispanic Acceptance Rate

```{r}
# Initialize an empty data frame to store results
hispanic_acceptance <- data.frame(Hispanic_Status = character(), Acceptance_Percent = numeric(), stringsAsFactors = FALSE)

# Define Hispanic response categories
hispanic_responses <- c("Hispanic or Latino/Latina/Latinx", 
                         "I prefer not to answer", 
                         "Not Provided", 
                         "Not Hispanic or Latino/Latina/Latinx")

# Loop over each Hispanic response and calculate acceptance percentage
for (hispanic_label in hispanic_responses) {  

  # Filter for total applications matching this Hispanic response
  total_apps <- data_clean_indiv %>%
    filter(hispanic == hispanic_label) %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Filter for accepted applications matching this Hispanic response
  accepted_apps <- data_clean_indiv %>%
    filter(hispanic == hispanic_label & Application.Status2 == "Accepted") %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Calculate acceptance rate
  acceptance_rate <- ifelse(total_apps > 0, (accepted_apps / total_apps) * 100, NA)

  # Append to results data frame
  hispanic_acceptance <- rbind(hispanic_acceptance, data.frame(Hispanic_Status = hispanic_label, Acceptance_Percent = round(acceptance_rate, 2)))
}

# Display the results as a table
hispanic_acceptance %>%
  kable(col.names = c("Hispanic Status", "Acceptance Rate (%)"),
        caption = "Acceptance Rate by Hispanic Status")

```

### LGBTQIA

```{r}
unique_name_lgbtqia_count <- data_clean %>%
  distinct(Name.Key, LGBTQIA) %>%
  count()

# Step 1: Define possible responses
lgbtqia_responses <- c("I prefer not to answer", 
                       "No, I do NOT identify as part of the LGBTQIA community", 
                       "Yes, I identify as part of the LGBTQIA community")

#filter out "Not Provided"
data_clean_indiv <- data_clean_indiv %>%
  filter(Applicant.Race != "Not Provided")

# Step 2: Count the occurrences of each response
lgbtqia_counts <- data_clean_indiv %>%
  filter(LGBTQIA %in% lgbtqia_responses) %>%
  group_by(LGBTQIA) %>%
  summarise(Count = n())

# Step 3: Calculate the percentage of LGBTQIA (excluding "I prefer not to answer" and "Not Provided")
lgbtqia_percentage <- data_clean_indiv %>%
  filter(LGBTQIA %in% c("Yes, I identify as part of the LGBTQIA community", "No, I do NOT identify as part of the LGBTQIA community")) %>%
  summarise(
    LGBTQIA_Yes_Count = sum(LGBTQIA == "Yes, I identify as part of the LGBTQIA community"),
    Total_Provided = n(),
    Percent_LGBTQIA = (LGBTQIA_Yes_Count / Total_Provided) * 100
  ) %>%
  pull(Percent_LGBTQIA)

# Step 4: Add the percentage LGBTQIA row to the dataset
lgbtqia_counts <- rbind(lgbtqia_counts, data.frame(LGBTQIA = "% LGBTQIA (of those who provided response)", Count = round(lgbtqia_percentage, 2)))

# Step 5: Reorder the rows as requested
lgbtqia_counts <- lgbtqia_counts %>%
  arrange(factor(LGBTQIA, levels = c("Not Provided", "I prefer not to answer", "No, I do NOT identify as part of the LGBTQIA community", "Yes, I identify as part of the LGBTQIA community", "% LGBTQIA (of those who provided response)")))

# Step 6: Format the Count column: no decimals for counts, two decimals for Percent_LGBTQIA
lgbtqia_counts <- lgbtqia_counts %>%
  mutate(Count = ifelse(LGBTQIA == "% LGBTQIA (of those who provided response)", 
                        format(Count, nsmall = 2), 
                        as.character(Count)))  # Keep other counts as whole numbers

# Use kable to display the counts table in a more presentable format
lgbtqia_counts %>%
  kable(
    caption = "LGBTQIA Counts and Percentage Breakdown",
    col.names = c("LGBTQIA Status", "Count"),
    align = c("l", "r"),
    format = "markdown"  # Or "html" for HTML output
  )

# Step 1: Calculate total counts and percentages
lgbtqia_counts <- data_clean_indiv %>%
  filter(LGBTQIA %in% c("I prefer not to answer", 
                        "No, I do NOT identify as part of the LGBTQIA community", 
                        "Yes, I identify as part of the LGBTQIA community")) %>%
  count(LGBTQIA) %>%
  mutate(Percent = (n / sum(n)) * 100)  # Calculate percent of total

# Step 2: Create the bar chart with updated layout
ggplot(lgbtqia_counts, aes(x = fct_relevel(LGBTQIA, 
                                           "I prefer not to answer", 
                                           "No, I do NOT identify as part of the LGBTQIA community", 
                                           "Yes, I identify as part of the LGBTQIA community"), 
                           y = Percent)) +  
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Percent)), vjust = -0.5) +  # Show % above bars
  geom_text(aes(label = n, y = -5), vjust = 0.4) +  # Raw count below x-axis labels
  labs(title = "LGBTQIA Identity Distribution of Sole Proprietor/Individual Applicants", 
       x = "LGBTQIA Identity", 
       y = "Percent of Responses") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     limits = c(-10, 100),  # Extends slightly below 0 for count visibility
                     expand = expansion(mult = c(0, 0.1))) +  # Prevents excess space
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),  # Adjust x-axis labels
    axis.title.x = element_text(margin = margin(t = 5)),  # Adjust x-axis title position
    plot.margin = margin(t = 10, r = 5, b = 10, l = 50)  # Expand left margin to fit first label
  )



```

LGBTQIA Acceptance

```{r}
# Initialize an empty data frame to store results
lgbtqia_acceptance <- data.frame(LGBTQIA_Status = character(), Acceptance_Percent = numeric(), stringsAsFactors = FALSE)

# Define LGBTQIA response categories
lgbtqia_responses <- c("I prefer not to answer", 
                        "Not Provided", 
                        "No, I do NOT identify as part of the LGBTQIA community", 
                        "Yes, I identify as part of the LGBTQIA community")

# Loop over each LGBTQIA response and calculate acceptance percentage
for (lgbtqia_label in lgbtqia_responses) {  

  # Filter for total applications matching this LGBTQIA response
  total_apps <- data_clean_indiv %>%
    filter(LGBTQIA == lgbtqia_label) %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Filter for accepted applications matching this LGBTQIA response
  accepted_apps <- data_clean_indiv %>%
    filter(LGBTQIA == lgbtqia_label & Application.Status2 == "Accepted") %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Calculate acceptance rate
  acceptance_rate <- ifelse(total_apps > 0, (accepted_apps / total_apps) * 100, NA)

  # Append to results data frame
  lgbtqia_acceptance <- rbind(lgbtqia_acceptance, data.frame(LGBTQIA_Status = lgbtqia_label, Acceptance_Percent = round(acceptance_rate, 2)))
}

# Display the results as a table
lgbtqia_acceptance %>%
  kable(col.names = c("LGBTQIA Status", "Acceptance Rate (%)"),
        caption = "Acceptance Rate by LGBTQIA Status")


```

```{r}
# Step 1: Create a summary table by grouping on LGBTQIA and Application.Status2
lgbtqia_status_summary <- data_clean_indiv %>%
  group_by(LGBTQIA, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 2: Pivot the table to create a crosstable
lgbtqia_status_pivot <- lgbtqia_status_summary %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Step 3: Calculate total applications per LGBTQIA category
lgbtqia_status_pivot <- lgbtqia_status_pivot %>%
  mutate(Total = rowSums(select(., -LGBTQIA)))

# Step 4: Calculate percentages for each status
lgbtqia_status_percent <- lgbtqia_status_pivot %>%
  mutate(
    `Accepted (%)` = (Accepted / Total) * 100,
    `In-Progress (%)` = (`In-Progress` / Total) * 100,
    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / Total) * 100,
    `Total (%)` = 100  # Since total column should always be 100%
  ) 

# Step 5: Ensure rounding and percentage formatting
lgbtqia_status_percent <- lgbtqia_status_percent %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%  # Round to 1 decimal
  mutate(across(ends_with("(%)"), ~ paste0(.x, "%")))  # Append "%" to percentage columns

# Step 6: Manually create the Totals Row
totals_row <- lgbtqia_status_pivot %>%
  summarise(across(-LGBTQIA, sum)) %>%
  mutate(LGBTQIA = "Total")

# Step 6.1: Append "100.0%" manually to all percentage columns
for (col in c("Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)", "Total (%)")) {
  totals_row[[col]] <- "100.0%"  # Explicitly set to "100.0%" for all status columns
}

# Step 7: Append the Totals Row to the dataset
lgbtqia_status_final <- bind_rows(lgbtqia_status_percent, totals_row)

# Step 8: Define correct ordering for LGBTQIA response categories
lgbtqia_responses <- c(
  "Yes, I identify as part of the LGBTQIA community", 
  "No, I do NOT identify as part of the LGBTQIA community", 
  "I prefer not to answer", 
  "Not Provided",
  "Total"
)

# Ensure proper order
lgbtqia_status_final <- lgbtqia_status_final %>%
  mutate(LGBTQIA = factor(LGBTQIA, levels = lgbtqia_responses)) %>%
  arrange(LGBTQIA)

# Step 9: Reorder columns to interleave count and percentage
lgbtqia_status_final <- lgbtqia_status_final %>%
  select(LGBTQIA, 
         Accepted, `Accepted (%)`, 
         `In-Progress`, `In-Progress (%)`, 
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`, 
         Total, `Total (%)`)  # Ensure alternating columns

# Step 10: Format and Display the Table
lgbtqia_status_final %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), 
    col.names = c("LGBTQIA Status", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Total", "Percent"),
    caption = "Application Status by LGBTQIA Status"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = FALSE, 
                position = "left", 
                font_size = 9,  
                latex_options = "scale_down") %>%
  add_header_above(c(" " = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%
  column_spec(1, width = "20em", extra_css = "min-width: 20em; max-width: 20em; text-align: left;") %>%  # Forces wider first column
  column_spec(2:9, width = "4em", extra_css = "min-width: 4em; max-width: 4em; text-align: right;") %>%  # Keeps other columns narrow
  row_spec(nrow(lgbtqia_status_final), bold = TRUE)  # Bold the totals row

```

### Age

```{r}

# Create the bar chart for Age
ggplot(data_clean_indiv, aes(x = age)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +  # Add count labels at the top of each bar
  labs(title = "Count of Applications by Age Group", 
       x = "Age Group", 
       y = "Count of Applications") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +  # Expand by 10% at the top
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 1: Calculate total counts and percentages
age_counts <- data_clean_indiv %>%
  count(age) %>%
  mutate(Percent = (n / sum(n)) * 100)  # Calculate percent of total

# Step 2: Create the bar chart with updated layout
ggplot(age_counts, aes(x = age, y = Percent)) +  
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Percent)), vjust = -0.5) +  # Show % above bars
  geom_text(aes(label = n, y = 0), vjust = 1.2) +  # Raw count below x-axis labels
  labs(title = "Count of Applications by Age Group", 
       x = "Age Group", 
       y = "Percent of Applications") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     expand = expansion(mult = c(0.2, 0.5), add = c(0, 0))) +  # Prevents negative y-axis
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1.015),  # Lower x-axis labels
    axis.title.x = element_text(margin = margin(t = 5)),  # Adjust x-axis title position
    plot.margin = margin(t = 10, r = 5, b = 10, l = 55)  # Slightly increase left margin
  )


```

Age Acceptance

```{r}
# Initialize an empty data frame to store results
age_acceptance <- data.frame(Age_Group = character(), Acceptance_Percent = numeric(), stringsAsFactors = FALSE)

# Get unique age groups from the dataset
age_groups <- unique(data_clean_indiv$age)

# Loop over each age group and calculate acceptance percentage
for (age_label in age_groups) {  

  # Filter for total applications matching this age group
  total_apps <- data_clean_indiv %>%
    filter(age == age_label) %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Filter for accepted applications matching this age group
  accepted_apps <- data_clean_indiv %>%
    filter(age == age_label & Application.Status2 == "Accepted") %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Calculate acceptance rate
  acceptance_rate <- ifelse(total_apps > 0, (accepted_apps / total_apps) * 100, NA)

  # Append to results data frame
  age_acceptance <- rbind(age_acceptance, data.frame(Age_Group = age_label, Acceptance_Percent = round(acceptance_rate, 2)))
}

# Define the desired order for Age_Group
age_levels <- c("18 - 24", "25 - 34", "35 - 44", "45 - 54", "55 - 64", "65 and over", "I prefer not to answer", "Not Provided")

# Convert Age_Group to a factor with the defined order
age_acceptance <- age_acceptance %>%
  mutate(Age_Group = factor(Age_Group, levels = age_levels)) %>%
  arrange(Age_Group)

# Display the reordered table
age_acceptance %>%
  kable(col.names = c("Age Group", "Acceptance Rate (%)"),
        caption = "Acceptance Rate by Age Group")


```

```{r}
# Step 1: Create a summary table by grouping on Age and Application.Status2
age_status_summary <- data_clean_indiv %>%
  group_by(age, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 2: Pivot the table to create a crosstable
age_status_pivot <- age_status_summary %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Step 3: Calculate total applications per age category
age_status_pivot <- age_status_pivot %>%
  mutate(Total = rowSums(select(., -age)))

# Step 4: Calculate percentages for each status
age_status_percent <- age_status_pivot %>%
  mutate(
    `Accepted (%)` = (Accepted / Total) * 100,
    `In-Progress (%)` = (`In-Progress` / Total) * 100,
    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / Total) * 100,
    `Total (%)` = 100  # Since total column should always be 100%
  ) 

# Step 5: Ensure rounding and percentage formatting
age_status_percent <- age_status_percent %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%  # Round to 1 decimal
  mutate(across(ends_with("(%)"), ~ paste0(.x, "%")))  # Append "%" to percentage columns

# Step 6: Manually create the Totals Row
totals_row <- age_status_pivot %>%
  summarise(across(-age, sum)) %>%
  mutate(age = "Total")

# Step 6.1: Append "100.0%" manually to all percentage columns
for (col in c("Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)", "Total (%)")) {
  totals_row[[col]] <- "100.0%"  # Explicitly set to "100.0%" for all status columns
}

# Step 7: Append the Totals Row to the dataset
age_status_final <- bind_rows(age_status_percent, totals_row)

# Step 8: Define correct ordering for Age categories
age_levels <- c(
  "18 - 24", 
  "25 - 34", 
  "35 - 44", 
  "45 - 54", 
  "55 - 64", 
  "65 and over", 
  "I prefer not to answer",
  "Not Provided",
  "Total"
)

# Ensure proper order
age_status_final <- age_status_final %>%
  mutate(age = factor(age, levels = age_levels)) %>%
  arrange(age)

# Step 9: Reorder columns to interleave count and percentage
age_status_final <- age_status_final %>%
  select(age, 
         Accepted, `Accepted (%)`, 
         `In-Progress`, `In-Progress (%)`, 
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`, 
         Total, `Total (%)`)  # Ensure alternating columns

# Step 10: Format and Display the Table
age_status_final %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), 
    col.names = c("Age Group", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Total", "Percent"),
    caption = "Application Status by Age Group"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = FALSE, 
                position = "left", 
                font_size = 12,  # Readable but compact
                latex_options = "scale_down") %>%
  add_header_above(c(" " = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%
  column_spec(1, width = "20em") %>%  # First column does not need excessive width
  column_spec(2:9, width = "7em") %>%  # Keeps other columns compact
  row_spec(nrow(age_status_final), bold = TRUE)  # Bold the totals row

```

### Income

```{r}
# Define the correct order for income levels
income_order <- c("Less than $25,000", 
                  "$25,000 to $49,999", 
                  "$50,000 to $74,999", 
                  "$75,000 to $99,999", 
                  "$100,000 to $124,999", 
                  "$125,000 to $149,999", 
                  "$150,000 or more")

# Reorder income levels
data_clean_indiv <- data_clean_indiv %>%
  mutate(income = factor(income, levels = c(income_order, "Not Provided"))) # Reorder levels

# Create the bar chart for income
ggplot(data_clean_indiv, aes(x = income)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +  # Add count labels at the top of each bar
  labs(
    title = "Count of Applications by Income Level", 
    x = "Income Level", 
    y = "Count of Applications"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +  # Expand by 10% at the top
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 1: Calculate total counts and percentages
income_counts <- data_clean_indiv %>%
  count(income) %>%
  mutate(Percent = (n / sum(n)) * 100)  # Calculate percent of total

# Step 2: Create the bar chart with updated layout
ggplot(income_counts, aes(x = income, y = Percent)) +  # Keep income order as-is
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Percent)), vjust = -0.5) +  # Show % above bars
  geom_text(aes(label = n, y = 0), vjust = 1.2) +  # Raw count below x-axis labels
  labs(
    title = "Count of Applications by Income Level", 
    x = "Income Level", 
    y = "Percent of Applications"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     expand = expansion(mult = c(0.2, 0.5), add = c(0, 0))) +  # Prevents negative y-axis
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1.015),  # Lower x-axis labels
    axis.title.x = element_text(margin = margin(t = 5)),  # Adjust x-axis title position
    plot.margin = margin(t = 10, r = 5, b = 10, l = 55)  # Slightly increase left margin
  )


```

Income Acceptance

```{r}
# Initialize an empty data frame to store results
income_acceptance <- data.frame(Income_Level = character(), Acceptance_Percent = numeric(), stringsAsFactors = FALSE)

# Define ordered income categories
income_levels <- c("Less than $25,000", 
                   "$25,000 to $49,999", 
                   "$50,000 to $74,999", 
                   "$75,000 to $99,999", 
                   "$100,000 to $124,999", 
                   "$125,000 to $149,999", 
                   "$150,000 or more", 
                   "Not Provided")

# Loop over each income level and calculate acceptance percentage
for (income_label in income_levels) {  

  # Filter for total applications matching this income level
  total_apps <- data_clean_indiv %>%
    filter(income == income_label) %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Filter for accepted applications matching this income level
  accepted_apps <- data_clean_indiv %>%
    filter(income == income_label & Application.Status2 == "Accepted") %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Calculate acceptance rate
  acceptance_rate <- ifelse(total_apps > 0, (accepted_apps / total_apps) * 100, NA)

  # Append to results data frame
  income_acceptance <- rbind(income_acceptance, data.frame(Income_Level = income_label, Acceptance_Percent = round(acceptance_rate, 2)))
}

# Convert Income_Level to a factor and reorder
income_acceptance <- income_acceptance %>%
  mutate(Income_Level = factor(Income_Level, levels = income_levels)) %>%
  arrange(Income_Level)

# Display the table
income_acceptance %>%
  kable(col.names = c("Income Level", "Acceptance Rate (%)"),
        caption = "Acceptance Rate by Income Level")

```

```{r}
# Step 1: Ensure "Not Provided" for Missing Income
data_clean_indiv <- data_clean_indiv %>%
  mutate(income = if_else(is.na(income) | income == "", "Not Provided", income))

# Step 2: Create a summary table by grouping on Income and Application.Status2
income_status_summary <- data_clean_indiv %>%
  group_by(income, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 3: Pivot the table to create a crosstable
income_status_pivot <- income_status_summary %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Step 4: Calculate total applications per income level
income_status_pivot <- income_status_pivot %>%
  mutate(Total = rowSums(select(., -income)))

# Step 5: Calculate percentages for each status
income_status_percent <- income_status_pivot %>%
  mutate(
    `Accepted (%)` = (Accepted / Total) * 100,
    `In-Progress (%)` = (`In-Progress` / Total) * 100,
    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / Total) * 100,
    `Total (%)` = 100  # Since total column should always be 100%
  ) 

# Step 6: Ensure rounding and percentage formatting
income_status_percent <- income_status_percent %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%  # Round to 1 decimal
  mutate(across(ends_with("(%)"), ~ paste0(.x, "%")))  # Append "%" to percentage columns

# Step 7: Manually create the Totals Row
totals_row <- income_status_pivot %>%
  summarise(across(-income, sum)) %>%
  mutate(income = "Total")

# Step 7.1: Append "100.0%" manually to all percentage columns
for (col in c("Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)", "Total (%)")) {
  totals_row[[col]] <- "100.0%"  # Explicitly set to "100.0%" for all status columns
}

# Step 8: Append the Totals Row to the dataset
income_status_final <- bind_rows(income_status_percent, totals_row)

# Step 9: Define correct ordering for Income categories
income_levels <- c(
  "Less than $25,000", 
  "$25,000 to $49,999", 
  "$50,000 to $74,999", 
  "$75,000 to $99,999", 
  "$100,000 to $124,999", 
  "$125,000 to $149,999", 
  "$150,000 or more", 
  "Not Provided",  # Now explicitly before Total
  "Total"
)

# Ensure proper order
income_status_final <- income_status_final %>%
  mutate(income = factor(income, levels = income_levels)) %>%
  arrange(income)

# Step 10: Reorder columns to interleave count and percentage
income_status_final <- income_status_final %>%
  select(income, 
         Accepted, `Accepted (%)`, 
         `In-Progress`, `In-Progress (%)`, 
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`, 
         Total, `Total (%)`)  # Ensure alternating columns

# Step 11: Format and Display the Table
income_status_final %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), 
    col.names = c("Income Level", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Total", "Percent"),
    caption = "Application Status by Income Level"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = FALSE, 
                position = "left", 
                font_size = 12,  # Readable font size
                latex_options = "scale_down") %>%
  add_header_above(c(" " = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%
  column_spec(1, width = "25em") %>%  # First column set to at least 25em
  column_spec(2:9, width = "6em") %>%  # Other columns set to 6em
  row_spec(nrow(income_status_final) - 1) %>%  # Bold "Not Provided"
  row_spec(nrow(income_status_final), bold = TRUE)  # Bold the totals row

```

### Neighborhood Connection

```{r}
# Reorder the levels of the connection2 variable
connection_order <- c("none", "<1 year", "1-5 years", "6-10 years", "11-20 years", "20+ years", "Not Provided")
sole_prop_data <- sole_prop_data %>%
  mutate(connection2 = factor(connection2, levels = connection_order))

# Create the bar chart with reordered connection2 variable
ggplot(sole_prop_data, aes(x = connection2)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +  # Add count labels at the top of each bar
  labs(title = "Count of Applications by Connection", 
       x = "Connection", 
       y = "Count of Applications") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +  # Expand by 10% at the top
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 1: Calculate total counts and percentages
connection_counts <- sole_prop_data %>%
  count(connection2) %>%
  mutate(Percent = (n / sum(n)) * 100)  # Calculate percent of total

# Step 2: Create the bar chart with updated layout
ggplot(connection_counts, aes(x = connection2, y = Percent)) +  # Keep connection order as-is
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Percent)), vjust = -0.5) +  # Show % above bars
  geom_text(aes(label = n, y = 0), vjust = 1.2) +  # Raw count below x-axis labels
  labs(title = "Count of Applications by Connection", 
       x = "Connection", 
       y = "Percent of Applications") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     expand = expansion(mult = c(0.2, 0.5), add = c(0, 0))) +  # Prevents negative y-axis
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1.015),  # Lower x-axis labels
    axis.title.x = element_text(margin = margin(t = 5)),  # Adjust x-axis title position
    plot.margin = margin(t = 10, r = 5, b = 10, l = 55)  # Slightly increase left margin
  )


```

Connection Acceptance

```{r}
# Initialize an empty data frame to store results
connection_acceptance <- data.frame(Connection = character(), Acceptance_Percent = numeric(), stringsAsFactors = FALSE)

# Define ordered connection categories
connection_levels <- c("none", "<1 year", "1-5 years", "6-10 years", "11-20 years", "20+ years", "Not Provided")

# Loop over each connection level and calculate acceptance percentage
for (connection_label in connection_levels) {  

  # Filter for total applications matching this connection level
  total_apps <- data_clean_indiv %>%
    filter(connection2 == connection_label) %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Filter for accepted applications matching this connection level
  accepted_apps <- data_clean_indiv %>%
    filter(connection2 == connection_label & Application.Status2 == "Accepted") %>%
    summarise(Count = n()) %>%
    pull(Count)

  # Calculate acceptance rate
  acceptance_rate <- ifelse(total_apps > 0, (accepted_apps / total_apps) * 100, NA)

  # Append to results data frame
  connection_acceptance <- rbind(connection_acceptance, data.frame(Connection = connection_label, Acceptance_Percent = round(acceptance_rate, 2)))
}

# Convert Connection to a factor and reorder
connection_acceptance <- connection_acceptance %>%
  mutate(Connection = factor(Connection, levels = connection_levels)) %>%
  arrange(Connection)

# Display the table
connection_acceptance %>%
  kable(col.names = c("Neighborhood Connection", "Acceptance Rate (%)"),
        caption = "Acceptance Rate by Neighborhood Connection")

```

```{r}
# Step 1: Create a summary table by grouping on Connection and Application.Status2
connection_status_summary <- sole_prop_data %>%
  group_by(connection2, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 2: Pivot the table to create a crosstable
connection_status_pivot <- connection_status_summary %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Step 3: Calculate total applications per connection level
connection_status_pivot <- connection_status_pivot %>%
  mutate(Total = rowSums(select(., -connection2)))

# Step 4: Calculate percentages for each status
connection_status_percent <- connection_status_pivot %>%
  mutate(
    `Accepted (%)` = (Accepted / Total) * 100,
    `In-Progress (%)` = (`In-Progress` / Total) * 100,
    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / Total) * 100,
    `Total (%)` = 100  # Since total column should always be 100%
  ) 

# Step 5: Ensure rounding and percentage formatting
connection_status_percent <- connection_status_percent %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%  # Round to 1 decimal
  mutate(across(ends_with("(%)"), ~ paste0(.x, "%")))  # Append "%" to percentage columns

# Step 6: Manually create the Totals Row
totals_row <- connection_status_pivot %>%
  summarise(across(-connection2, sum)) %>%
  mutate(connection2 = "Total")

# Step 6.1: Append "100.0%" manually to all percentage columns
for (col in c("Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)", "Total (%)")) {
  totals_row[[col]] <- "100.0%"  # Explicitly set to "100.0%" for all status columns
}

# Step 7: Append the Totals Row to the dataset
connection_status_final <- bind_rows(connection_status_percent, totals_row)

# Step 8: Define correct ordering for Connection categories
connection_levels <- c(
  "none", 
  "<1 year", 
  "1-5 years", 
  "6-10 years", 
  "11-20 years", 
  "20+ years", 
  "Not Provided",
  "Total"
)

# Ensure proper order
connection_status_final <- connection_status_final %>%
  mutate(connection2 = factor(connection2, levels = connection_levels)) %>%
  arrange(connection2)

# Step 9: Reorder columns to interleave count and percentage
connection_status_final <- connection_status_final %>%
  select(connection2, 
         Accepted, `Accepted (%)`, 
         `In-Progress`, `In-Progress (%)`, 
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`, 
         Total, `Total (%)`)  # Ensure alternating columns

# Step 10: Format and Display the Table
connection_status_final %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), 
    col.names = c("Neighborhood Connection", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Count", "Percent", 
                  "Total", "Percent"),
    caption = "Application Status by Neighborhood Connection"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = FALSE, 
                position = "left", 
                font_size = 12,  # Readable font size
                latex_options = "scale_down") %>%
  add_header_above(c(" " = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%
  column_spec(1, width = "25em") %>%  # First column set to at least 25em
  column_spec(2:9, width = "6em") %>%  # Other columns set to 6em
  row_spec(nrow(connection_status_final), bold = TRUE)  # Bold the totals row

```

# Application Status by community Area

```{r}
data_clean_comm_area <- data_clean %>% 
  filter(!is.na(lon)) 

# make dataframe into spatial feature 
comm_area_apps_sf <- st_as_sf(x= data_clean_comm_area, coords = c("lon", "lat"), crs = 4326)

comm_area_boundaries <- read_sf("geo_export_a446ed1e-4dd2-4beb-b7b6-6fb916b4bdde.shp")
comm_area_boundaries <- st_transform(comm_area_boundaries, crs = 3435)

#change the application layer CRS to equal 3435
comm_area_apps_sf <- st_transform(comm_area_apps_sf, crs = st_crs(comm_area_boundaries))

# Join polygon features to points
comm_area_data <- st_join(comm_area_apps_sf, comm_area_boundaries, join = st_intersects)

comm_area_table <- comm_area_data %>%
  group_by(community) %>%
  summarise(Unique_Property_PINs = n_distinct(Property.PIN)) %>%
  arrange(desc(Unique_Property_PINs))

comm_area_data$Application.Status2 <- comm_area_data$Application.Status  # Step 1: Create the field

comm_area_data$Application.Status2 <- ifelse(comm_area_data$Application.Status %in% c("Withdrawn", "Declined"),
                                         "Declined & Withdrawn",
                                         comm_area_data$Application.Status2)


comm_area_data <- st_drop_geometry(comm_area_data)

comm_area_status_count_table <- comm_area_data %>%
  group_by(community, Application.Status2) %>%
  summarise(Count = n()) %>%
  ungroup()

comm_area_status_pivot_count_table <- comm_area_status_count_table %>%
  dplyr::select(community, Application.Status2, Count) %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

comm_area_status_pivot_count_table <- comm_area_status_pivot_count_table %>%
  dplyr::select(community, "Accepted", "In-Progress", "Declined & Withdrawn")


comm_area_status_percent_table <- comm_area_data %>%
  group_by(community, Application.Status2) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  ungroup()

total_count_by_comm_area <- comm_area_status_percent_table %>%
  group_by(community) %>%
  summarize(Total_Count = sum(Count, na.rm = TRUE))

comm_area_status_pivot_table <- comm_area_status_percent_table %>%
  dplyr::select(community, Application.Status2, Percentage) %>%
  pivot_wider(names_from = Application.Status2, values_from = Percentage, values_fill = 0) %>%
  left_join(total_count_by_comm_area, by = "community")

comm_area_status_pivot_table <- comm_area_status_pivot_table %>%
  dplyr::select(community, Total_Count, "Accepted", "In-Progress", "Declined & Withdrawn")

comm_area_status_pivot_table <- comm_area_status_pivot_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))

######## CREATE DEMAND SCORE 

#ensure universe is clean
# filtered_data <- data_clean %>%
#   filter(!is.na(Property_ZipCode) & Property_ZipCode != 0)

# Get unique Property.PIN values
unique_lots_commarea <- comm_area_data %>%
  distinct(Property.PIN, community)

# Count unique lots by Community Area
commarea_lot_counts <- unique_lots_commarea %>%
  group_by(community) %>%
  summarise(Unique.Lots.by.Commarea = n(), .groups = "drop")

# Check the total record count
nrow(unique_lots_commarea)  # Should returned 1100 records

# Count unique applications by Community Area
commarea_app_counts <- comm_area_data %>%
  filter(!is.na(Property_ZipCode) & Property_ZipCode != 0) %>%
  group_by(community) %>%
  summarise(Property.Applications.by.Commarea = n(), .groups = "drop")

# Join the two values into one dataset
commarea_summary <- commarea_lot_counts %>%
  full_join(commarea_app_counts, by = c("community" = "community")) 

# Calculate "Demand" 
commarea_summary <- commarea_summary %>%
  mutate(Commarea.Demand.Score = Property.Applications.by.Commarea / Unique.Lots.by.Commarea)

commarea_summary <- commarea_summary %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%
  select(community,Commarea.Demand.Score)

######## JOIN DEMAND SCORE TO TABLE

comm_area_status_pivot_table <- left_join(comm_area_status_pivot_table, commarea_summary, by = "community")

#Reorder the columns 
comm_area_status_pivot_table <- comm_area_status_pivot_table %>%
  dplyr::select(community, Total_Count, Commarea.Demand.Score, "Accepted", "In-Progress", "Declined & Withdrawn")

######## CREATE TABLE

comm_area_status_pivot_table %>%
  arrange(desc(Total_Count)) %>%
  filter(Total_Count >= 10) %>%
  kable(col.names = c("Community Area", "Count of Applications", "Demand Score", "Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)"),
        caption = "Percent of Application Status by Community Area")

```

Matt Request Table Edit  Entity Type

```{r}
# Step 1: Calculate counts for each Application.Status2 by Purchasing.Entity.Type
status_count_table <- data_clean %>%
  group_by(Purchasing.Entity.Type, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 2: Pivot the table to wide format (statuses as columns)
status_count_pivot <- status_count_table %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Step 3: Calculate total applications per entity type
status_count_pivot <- status_count_pivot %>%
  mutate(Total = rowSums(select(., -Purchasing.Entity.Type)))

# Step 4: Calculate percentages for each status
status_percent_table <- status_count_pivot %>%
  mutate(
    `Accepted (%)` = (Accepted / Total) * 100,
    `In-Progress (%)` = (`In-Progress` / Total) * 100,
    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / Total) * 100
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))  # Round to 1 decimal

#Select only necessary columns from the percent table before joining
status_percent_table <- status_percent_table %>%
  select(Purchasing.Entity.Type, `Accepted (%)`, `In-Progress (%)`, `Declined & Withdrawn (%)`)

library(scales)  # Needed for comma formatting

# Step 5: Merge Count and Percent into one final table (avoiding .x and .y naming)
final_table <- status_count_pivot %>%
  left_join(status_percent_table, by = "Purchasing.Entity.Type") %>%
  select(Purchasing.Entity.Type,
         Accepted, `Accepted (%)`,
         `In-Progress`, `In-Progress (%)`,
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`,
         Total)  # Ensure correct order

final_table <- final_table %>%
  mutate(`Total (%)` = 100.0)  # Since total is always 100%

# Apply formatting:
final_table <- final_table %>%
  mutate(
    across(c(Accepted, `In-Progress`, `Declined & Withdrawn`, Total),
           ~ formatC(.x, format = "f", big.mark = ",", digits = 0)),  # Comma formatting, no decimals
    across(c(`Accepted (%)`, `In-Progress (%)`, `Declined & Withdrawn (%)`, `Total (%)`),
           ~ paste0(formatC(.x, format = "f", digits = 1), "%"))  # Ensure 1 decimal place & append "%"
  )

# Create the formatted table with multi-line headers
final_table %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), col.names = NULL) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  add_header_above(c(" " = 1, "Count" = 1, "Percent" = 1,
                     "Count" = 1, "Percent" = 1,
                     "Count" = 1, "Percent" = 1,
                     "Count" = 1, "Percent" = 1)) %>%
  add_header_above(c("Entity Type" = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%
  column_spec(2:9, width = "7em", extra_css = "text-align: right;") %>%
  row_spec(0, align = "left")


# Step 5: Merge Count and Percent into one final table (avoiding .x and .y naming)
final_table <- status_count_pivot %>%
  left_join(status_percent_table, by = "Purchasing.Entity.Type") %>%
  select(Purchasing.Entity.Type,
         Accepted, `Accepted (%)`,
         `In-Progress`, `In-Progress (%)`,
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`,
         Total)  # Ensure correct order

final_table <- final_table %>%
  mutate(`Total (%)` = 100.0)  # Since total is always 100%

final_table <- final_table %>%
  mutate(across(c(`Accepted (%)`, `In-Progress (%)`, `Declined & Withdrawn (%)`, `Total (%)`),
                ~ paste0(.x, "%")))

# Create the formatted table with multi-line headers
final_table %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), col.names = NULL) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
    add_header_above(c(" " = 1, "Count" = 1, "Percent" = 1,
                              "Count" = 1, "Percent" = 1,
                              "Count" = 1, "Percent" = 1,
                              "Count" = 1, "Percent" = 1)) %>%
  add_header_above(c("Entity Type" = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%
  column_spec(2:9, width = "7em", extra_css = "text-align: right;") %>%
  row_spec(0, align = "left")
```

Matt Request Table Edit  Application Lot Type

```{r}

# Step 1: Calculate counts for each Application.Status2 by Application.Type

status_count_table <- data_clean %>%

  group_by(Application.Type, Application.Status2) %>%

  summarise(Count = n(), .groups = "drop")

 

# Step 2: Pivot the table to wide format (statuses as columns)

status_count_pivot <- status_count_table %>%

  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

 

# Step 3: Calculate total applications per Application Type

status_count_pivot <- status_count_pivot %>%

  mutate(Total = rowSums(select(., -Application.Type)))

 

# Step 4: Calculate percentages for each status

status_percent_table <- status_count_pivot %>%

  mutate(

    `Accepted (%)` = (Accepted / Total) * 100,

    `In-Progress (%)` = (`In-Progress` / Total) * 100,

    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / Total) * 100

  ) %>%

  mutate(across(where(is.numeric), ~ round(.x, 1)))  # Round to 1 decimal

 

# Step 5: Select only necessary columns from the percent table before joining

status_percent_table <- status_percent_table %>%

  select(Application.Type, `Accepted (%)`, `In-Progress (%)`, `Declined & Withdrawn (%)`)

 

# Step 6: Merge Count and Percent into one final table (avoiding .x and .y naming)

final_table <- status_count_pivot %>%

  left_join(status_percent_table, by = "Application.Type") %>%

  select(Application.Type,

         Accepted, `Accepted (%)`,

         `In-Progress`, `In-Progress (%)`,

         `Declined & Withdrawn`, `Declined & Withdrawn (%)`,

         Total)  # Ensure correct order

 

# Step 7: Add Total (%) column

final_table <- final_table %>%

  mutate(`Total (%)` = 100.0)  # Since total is always 100%

 

# Step 8: Format percentage columns to include "%" symbol

final_table <- final_table %>%

  mutate(across(c(`Accepted (%)`, `In-Progress (%)`, `Declined & Withdrawn (%)`, `Total (%)`),

                ~ paste0(.x, "%")))

 

# Step 9: Create the formatted table with multi-line headers

final_table %>%

  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c"), col.names = NULL) %>%

  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%

    add_header_above(c(" " = 1, "Count" = 1, "Percent" = 1,

                              "Count" = 1, "Percent" = 1,

                              "Count" = 1, "Percent" = 1,

                              "Count" = 1, "Percent" = 1)) %>%

  add_header_above(c("Application Type" = 1, "Accepted" = 2, "In-Progress" = 2, "Declined & Withdrawn" = 2, "Total" = 2)) %>%

  column_spec(2:9, width = "7em", extra_css = "text-align: right;") %>%

  row_spec(0, align = "left")

 

```

New Race Section

```{r}

data_clean_race <- data_clean_indiv %>%
  filter(Applicant.Race != "Not Provided")


#that gives us 815 application line items from sole proprietors who have provided a response on race
#long if statement to determine new Race & Ethnicity value

# Add the new column race_ethnicity
data_clean_race <- data_clean_race %>%
  mutate(race_ethnicity = case_when(

    # 1) Hispanic response
    hispanic == "Hispanic or Latino/Latina/Latinx" ~ "Hispanic or Latino (Any Race)",
    # 2) Both Hispanic & Race "Prefer not to answer"
    hispanic == "I prefer not to answer" & str_detect(Applicant.Race, "I prefer not to answer") ~ "I prefer not to answer",
    # 3) Hispanic "Prefer not to answer" & Race "I identify as a race that is not listed here"
    hispanic == "I prefer not to answer" & Applicant.Race == "I identify as a race that is not listed here" ~ "I identify as a race that is not listed here alone",
  # 4-6) Hispanic "Prefer not to answer" & Single Race
  hispanic == "I prefer not to answer" & Applicant.Race == "Black or African American" ~ "Black or African American alone (non-Hispanic or Latino)",
  hispanic == "I prefer not to answer" & Applicant.Race == "Native American or Alaskan Native" ~ "Native American or Alaskan Native alone (non-Hispanic or Latino)",
  hispanic == "I prefer not to answer" & Applicant.Race == "White or Caucasian" ~ "White or Caucasian alone (non-Hispanic or Latino)",
  # 7) Not Hispanic & Race contains "I prefer not to answer"
    hispanic == "Not Hispanic or Latino/Latina/Latinx" & str_detect(Applicant.Race, "I prefer not to answer") ~ "I prefer not to answer",
    # 8-12) Not Hispanic & Single Race
    hispanic == "Not Hispanic or Latino/Latina/Latinx" & Applicant.Race == "White or Caucasian" ~ "White or Caucasian alone (non-Hispanic or Latino)",
    hispanic == "Not Hispanic or Latino/Latina/Latinx" & Applicant.Race == "Black or African American" ~ "Black or African American alone (non-Hispanic or Latino)",
    hispanic == "Not Hispanic or Latino/Latina/Latinx" & Applicant.Race == "Asian or Pacific Islander" ~ "Asian or Pacific Islander alone (non-Hispanic or Latino)",
    hispanic == "Not Hispanic or Latino/Latina/Latinx" & Applicant.Race == "Native American or Alaskan Native" ~ "Native American or Alaskan Native alone (non-Hispanic or Latino)",
    hispanic == "Not Hispanic or Latino/Latina/Latinx" & Applicant.Race == "I identify as a race that is not listed here" ~ "I identify as a race that is not listed here alone",
    # 13) Everything else
    TRUE ~ "Two or more races (non-Hispanic or Latino)"
  ))

# Count occurrences of each race_ethnicity category

race_ethnicity_counts <- data_clean_race %>%
  count(race_ethnicity)

# Display the counts to verify
race_ethnicity_counts %>%
  kable(col.names = c("Race & Ethnicity", "Count"),
        caption = "Count of Race & Ethnicity Categories")
# Define race & ethnicity response categories (based on race_ethnicity field)
race_responses <- c("Black or African American alone (non-Hispanic or Latino)",
                    "Hispanic or Latino (Any Race)",
                    "White or Caucasian alone (non-Hispanic or Latino)",
                    "I identify as a race that is not listed here alone",
                    "Asian or Pacific Islander alone (non-Hispanic or Latino)",
                    "Two or more races (non-Hispanic or Latino)",
                    "Native American or Alaskan Native alone (non-Hispanic or Latino)",
                    "I prefer not to answer")

# Count occurrences directly from data_clean_race using race_ethnicity
race_counts <- data_clean_race %>%
  filter(race_ethnicity %in% race_responses) %>%  # Ensure only relevant categories
  count(race_ethnicity) %>%
  rename(Race = race_ethnicity, Count = n)
race_counts <- race_counts %>%
  mutate(Race = factor(Race, levels = race_responses))
 
# Generate a bar chart to visualize race counts
ggplot(race_counts, aes(x = Race, y = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5) +  # Add count labels above the bars
  labs(title = "Race & Ethnicity Distribution of Sole Proprietor/Individual Applicants",
       x = "Race & Ethnicity",
       y = "Count of Responses") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +  # Expand y-axis by 15% for label space
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Adjusted Race & Ethnicity to percent chart

```{r}
# Calculate total responses for percentage conversion
total_responses <- sum(race_counts$Count)

# Convert count to percentage and format for display
race_counts <- race_counts %>%
  mutate(Percent = (Count / total_responses) * 100,  # Convert to percentage
         Percent_Label = paste0(round(Percent, 1), "%"))  # Format percentage label

ggplot(race_counts, aes(x = Race, y = Percent)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Percent_Label), vjust = -0.5) +  # Keeps percent labels above bars
  geom_text(aes(label = Count, y = -5), vjust = 0.5) +  # Moves raw count closer to x-axis labels
  labs(title = "Race & Ethnicity Distribution of Sole Proprietor/Individual Applicants", 
       x = "Race & Ethnicity", 
       y = "Percent of Responses") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format(scale = 1), expand = expansion(mult = c(0.2, 0.5))) +  # Ensures proper top margin
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1.025),  # Keeps x-axis labels readable
    axis.title.x = element_text(margin = margin(t = 5)),  # Moves x-axis title up slightly to remove whitespace
    plot.margin = margin(t = 10, r = 5, b = 10, l = 50)  # Expands left margin so first label fits
  )

```

```{r}
# Step 1: Create a summary table by grouping on race_ethnicity and Application.Status2
race_status_summary <- data_clean_race %>%
  group_by(race_ethnicity, Application.Status2) %>%
  summarise(Count = n(), .groups = "drop")

# Step 2: Pivot the table to create a crosstable
race_status_pivot <- race_status_summary %>%
  pivot_wider(names_from = Application.Status2, values_from = Count, values_fill = 0)

# Step 3: Calculate total applications per race category
race_status_pivot <- race_status_pivot %>%
  mutate(Total = rowSums(select(., -race_ethnicity)))

# Step 4: Calculate percentages for each status
race_status_percent <- race_status_pivot %>%
  mutate(
    `Accepted (%)` = (Accepted / Total) * 100,
    `In-Progress (%)` = (`In-Progress` / Total) * 100,
    `Declined & Withdrawn (%)` = (`Declined & Withdrawn` / Total) * 100,
    `Total (%)` = 100  # Since total column should always be 100%
  ) 

# Step 5: Ensure rounding and percentage formatting
race_status_percent <- race_status_percent %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%  # Round to 1 decimal
  mutate(across(ends_with("(%)"), ~ paste0(.x, "%")))  # Append "%" to percentage columns

# Step 6: Manually create the Totals Row
totals_row <- race_status_pivot %>%
  summarise(across(-race_ethnicity, sum)) %>%
  mutate(race_ethnicity = "Total")

# Step 6.1: Append "100.0%" manually to all percentage columns
for (col in c("Accepted (%)", "In-Progress (%)", "Declined & Withdrawn (%)", "Total (%)")) {
  totals_row[[col]] <- "100.0%"  # Explicitly set to "100.0%" for all status columns
}

# Step 7: Append the Totals Row to the dataset
race_status_final <- bind_rows(race_status_percent, totals_row)

# Step 8: Define correct ordering for race categories
race_ethnicity_groups <- c(
  "Black or African American alone (non-Hispanic or Latino)", 
  "Hispanic or Latino (Any Race)", 
  "White or Caucasian alone (non-Hispanic or Latino)", 
  "I identify as a race that is not listed here alone", 
  "Asian or Pacific Islander alone (non-Hispanic or Latino)", 
  "Two or more races (non-Hispanic or Latino)", 
  "Native American or Alaskan Native alone (non-Hispanic or Latino)", 
  "I prefer not to answer",
  "Total"
)

# Ensure proper order
race_status_final <- race_status_final %>%
  mutate(race_ethnicity = factor(race_ethnicity, levels = race_ethnicity_groups)) %>%
  arrange(race_ethnicity)

# Step 9: Reorder columns to interleave count and percentage
race_status_final <- race_status_final %>%
  select(race_ethnicity, 
         Accepted, `Accepted (%)`, 
         `In-Progress`, `In-Progress (%)`, 
         `Declined & Withdrawn`, `Declined & Withdrawn (%)`, 
         Total, `Total (%)`)  # Ensure alternating columns

# Step 10: Format and Display the Table
race_status_final %>%
  kable(align = c("l", rep("c", 8)), 
        col.names = c("Race & Ethnicity", 
                      "Count", "Percent", 
                      "Count", "Percent", 
                      "Count", "Percent", 
                      "Total", "Percent"),
        caption = "Application Status by Race & Ethnicity",
        longtable = TRUE) %>%  # Helps in LaTeX/PDF output
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = FALSE, 
                position = "left", 
                font_size = 9,  # Forces smaller text to prevent stretching
                latex_options = "scale_down") %>%  # Scales down for PDF output
  add_header_above(c(" " = 1, 
                     "Accepted" = 2, 
                     "In-Progress" = 2, 
                     "Declined & Withdrawn" = 2, 
                     "Total" = 2)) %>%
  column_spec(1, width = "20em", extra_css = "min-width: 20em; max-width: 25em;") %>%  # Expands first column width in HTML
  column_spec(2:9, width = "4em", extra_css = "min-width: 4em; max-width: 6em;") %>%  # Keeps other columns narrower
  row_spec(nrow(race_status_final), bold = TRUE)

```

# Maps


```{r}
data_clean <- data_clean %>% rename(applicant_ZIP = Applicant.ZIP.Code)

# take all data and deduplicate based on name and applicant zip code.
zip_data <- data_clean %>%
  distinct(Name.Key, applicant_ZIP) %>%
  group_by(applicant_ZIP) %>%
  summarise(Count = n()) %>%
  filter(!is.na(applicant_ZIP) & applicant_ZIP != "")  # Filter out missing ZIPs


```

Exploring the home zip code of the applicants, there are 1,300 name:applicant zip combinations.

23 applicants provided a home ZIP Code from outside of Chicago.

```{r, eval=FALSE, warning=FALSE}
# Download ZIP code shapefiles for Cook County (includes Chicago)
zip_shapes <- zctas(cb = FALSE, year = 2010, state = "IL")

# Transform to correct CRS if necessary
zip_shapes <- st_transform(zip_shapes, crs = 3435)
st_write(zip_shapes, "zip_shapes.shp")
```

```{r}
zip_shapes <- read_sf("zip_shapes.shp")
# Step 1: Identify and summarize ZIPs outside of Illinois by performing an anti_join
outside_il_data <- zip_data %>%
  anti_join(zip_shapes, by = c("applicant_ZIP" = "ZCTA5CE10")) %>%
  summarise(Outside_IL_Count = sum(Count))

# Step 2: Filter out applicants with ZIPs outside Illinois
zip_data_il <- zip_data %>%
  semi_join(zip_shapes, by = c("applicant_ZIP" = "ZCTA5CE10"))

zip_shapes_merged <- zip_shapes %>%
  left_join(zip_data_il, by = c("ZCTA5CE10" = "applicant_ZIP"))  # Adjust column names if needed

# Fill NAs in Count with 0 (for ZIPs with no applicants)
zip_shapes_merged$Count[is.na(zip_shapes_merged$Count)] <- 0

```

# Figure 10: Lot Count by Zip Code as of February 11, 2025

```{r}
zip_shapes_chi <- read_sf("zip_shapes.shp")

# 
zip_data_prop <- data_clean %>%
  # Create unique combinations of property PINs and zip codes, so that properties only appear once
  distinct(Property.PIN, Property_ZipCode) %>%
  group_by(Property_ZipCode) %>%
  summarise(Count = n()) %>%
  filter(!is.na(Property_ZipCode) & Property_ZipCode != "")  # Filter out missing ZIPs
#there are 1,101 lots with valid ZIP codes

view_sum <- zip_data_prop %>%
  summarise(total = sum(Count, na.rm = TRUE))

# Convert Property_ZIPCode to character
zip_data_prop <- zip_data_prop %>%
  mutate(Property_ZipCode = as.character(Property_ZipCode))

# Filter out applicants with ZIPs outside Illinois
zip_data_il <- zip_data_prop %>%
  semi_join(zip_shapes_chi, by = c("Property_ZipCode" = "ZCTA5CE10"))
#still 1,126 lots with valid ZIP codes

zip_shapes_merged_chi <- zip_shapes_chi %>%
  left_join(zip_data_il, by = c("ZCTA5CE10" = "Property_ZipCode"))  # Adjust column names if needed

# Fill NAs in Count with 0 (for ZIPs with no applicants)
zip_shapes_merged_chi$Count[is.na(zip_shapes_merged_chi$Count)] <- 0

zip_shapes_merged_chi <- st_transform(zip_shapes_merged_chi, crs = st_crs(city_boundary))

#Spatial join
zips_chi_join <- zip_shapes_merged_chi[st_intersects(zip_shapes_merged_chi, city_boundary, sparse = FALSE), ]

zips_chi_join <- zips_chi_join %>%
  mutate(Count = replace_na(Count, 0))

zips_chi_join <- st_intersection(zips_chi_join, city_boundary)

tm_shape(zips_chi_join) +
  tm_polygons(fill = "Count", 
              fill.scale = tm_scale_intervals(n = 5, style = "jenks", values = "brewer.blues")) +
  tm_layout(legend.outside = TRUE, frame = FALSE) +
  tm_title("Lot Count by Zip Code")


# # Drop geometry
# zips_chi_join <- st_drop_geometry(zips_chi_join)
# 
# # Export to CSV
# write.csv(zips_chi_join, "Figure_10_March4.csv", row.names = FALSE)

```

# Figure 11: Count of Applicants by Residential ZIP Code as of February 11, 2025

```{r, message=FALSE, warning=FALSE}

# APPLICANT COUNT BY ZIP CODE

#remove zip codes with applicant counts of zero
# zip_shapes_merged_cleaned <- zip_shapes_merged %>%
#  filter(Count > 0)

#read in city boundary
city_boundary <- read_sf("city_boundary.shp")
city_boundary <- st_transform(city_boundary, crs = 3435)

#Align CRCS between the two files to join
zip_shapes_merged <- st_transform(zip_shapes_merged, crs = st_crs(city_boundary))
# 1,384 records in zip_shapes_merged 

# filter out zips that are mostly outside Chicago 
#zip_shapes_merged <- zip_shapes_merged %>%
 # filter()

#Spatial join
zips_chi_app <- zip_shapes_merged[st_intersects(zip_shapes_merged, city_boundary, sparse = FALSE), ]

zips_chi_app <- zips_chi_app %>%
  mutate(Count = replace_na(Count, 0))

zips_chi_app <- st_intersection(zips_chi_app, city_boundary)

tm_shape(zips_chi_app) +
  tm_polygons(fill = "Count", 
              fill.scale = tm_scale_intervals(n = 5, style = "jenks", values = "brewer.blues")) +
  tm_layout(legend.outside = TRUE, frame = FALSE) +
  tm_title("Applicant Count by Zip Code")


```

# Figure 12 Percent of applicants from within the same ZIP code

```{r}

# For zip codes that are the same for applicant and property, summarise applications (rows) by zip code and count the number of times that each zip appears in a new column called "local_count"
matched_zips <- data_clean %>%
  filter(applicant_ZIP == Property_ZipCode) %>% # filters rows where applicant_ZIP matches Property_ZipCode, keeping only records where the ZIP codes are the same.
  distinct(Name.Key, .keep_all = TRUE) %>% # Removes duplicate rows based on the Name.Key column.
  group_by(applicant_ZIP) %>%
  summarise(local_count = n()) # Counts the number of rows in each group (i.e., the number of unique Name.Key entries per applicant_ZIP).

# For ALL zip codes, summarise applicants by zip code of the property they applied to and count the number of applicants in a new column called "total_count"
total_zips <- data_clean %>%
  distinct(Name.Key, .keep_all = TRUE) %>%
  group_by(Property_ZipCode) %>%
  summarise(total_count = n())

total_zips <- total_zips %>%
  mutate(Property_ZipCode = as.character(Property_ZipCode))

joined_zips <- inner_join(total_zips,matched_zips, by = c("Property_ZipCode" = "applicant_ZIP"))

# Calculate the percentage of local applicants
joined_zips <- joined_zips %>%
  mutate(percent_local = (local_count / total_count) * 100)

percent_local_zips <- left_join(zips_chi_join, joined_zips, by = c("ZCTA5CE10" = "Property_ZipCode"))

percent_local_zips <- percent_local_zips %>%
  mutate(percent_local = replace_na(percent_local, 0))

percent_local_zips <- st_intersection(percent_local_zips, city_boundary)


tm_shape(percent_local_zips) +
  tm_polygons(fill = "percent_local", 
              fill.scale = tm_scale_intervals(n = 5, style = "jenks", values = "brewer.blues")) +
  tm_layout(legend.outside = TRUE, frame = FALSE) +
  tm_title("Percent of Applicants from within same ZIP Code")


```

# Figure 13: Percent of applications from within the same ZIP code for Individuals/Sole Proprietors only

```{r}

data_clean_individual <- data_clean %>%
  filter(Entity_Type_Group == "Sole Proprietor/Individual")

matched_zips <- data_clean %>%
  filter(applicant_ZIP == Property_ZipCode) %>% # filters rows where applicant_ZIP matches Property_ZipCode, keeping only records where the ZIP codes are the same.
  distinct(Name.Key, .keep_all = TRUE) %>% # Removes duplicate rows based on the Name.Key column.
  group_by(applicant_ZIP) %>%
  summarise(local_count = n()) # Counts the number of rows in each group (i.e., the number of unique Name.Key entries per applicant_ZIP).

# For ALL zip codes, summarise applicants by zip code of the property they applied to and count the number of applicants in a new column called "total_count"
total_zips <- data_clean %>%
  distinct(Name.Key, .keep_all = TRUE) %>%
  group_by(Property_ZipCode) %>%
  summarise(total_count = n())

total_zips <- total_zips %>%
  mutate(Property_ZipCode = as.character(Property_ZipCode))

joined_zips <- inner_join(total_zips,matched_zips, by = c("Property_ZipCode" = "applicant_ZIP"))

# Calculate the percentage of local applicants
joined_zips <- joined_zips %>%
  mutate(percent_local = (local_count / total_count) * 100)

percent_local_zips <- left_join(zips_chi_join, joined_zips, by = c("ZCTA5CE10" = "Property_ZipCode"))

percent_local_zips <- percent_local_zips %>%
  mutate(percent_local = replace_na(percent_local, 0))

percent_local_zips <- st_intersection(percent_local_zips, city_boundary)

tm_shape(percent_local_zips) +
  tm_polygons(fill = "percent_local", 
              fill.scale = tm_scale_intervals(n = 5, style = "jenks", values = "brewer.blues")) +
  tm_layout(legend.outside = TRUE, frame = FALSE) +
  tm_title("Percent of Applicants from within same ZIP Code for Sole Proprietors/Individuals")


```

All of the pale zips have low single digit applicant counts.
The darker the blue, the higher the concentration of applicant "home zip codes".

Property ZIP vs Applicant ZIP

```{r}

# Calculate the number of ZIP matches
match_count <- data_clean %>%
  filter(applicant_ZIP == Property_ZipCode) %>%
  nrow()

# Total number of rows
total_count <- nrow(data_clean)

# Calculate the percentage of matching rows
match_percentage <- (match_count / total_count) * 100

```

# Figure 16: Residential ZIP Code Locations of Applicants for Lots in 60612 Zip Code as of February 11, 2025

```{r}

shapefile_path <- "D:/GCI/CBB R/city_boundary.shp"
chicago_boundary <- st_read(shapefile_path)

# Calculate the number of non-ZIP matches
diff_zips <- data_clean %>%
  filter(applicant_ZIP != Property_ZipCode)
# 1,051 records found with non-matching zip-codes

#Break diff zips into two different dataframes
applicant_zips <- diff_zips
property_zips <- diff_zips

# Join zip shapefile to applicant_zips
applicat_zips_sf <- zip_shapes %>%
  inner_join(applicant_zips, by = c("ZCTA5CE10" = "applicant_ZIP"))
# 1,009 applicant with ZIP codes in IL

# Convert Property_ZIPCode to character
property_zips <- property_zips %>%
  mutate(Property_ZipCode = as.character(Property_ZipCode))

property_points_sf <- zip_shapes %>%
  inner_join(property_zips, by = c("ZCTA5CE10" = "Property_ZipCode"))

# Change zip polygons to centroid points
applicant_zips_points_sf <- st_centroid(applicat_zips_sf)
# applicant_zips_points_sf has 1,009 records
property_zips_points_sf <- st_centroid(property_points_sf)
# property_zips_points_sf has 828 records


# Transform applicant points to match Chicago CRS
applicant_zips_points_sf <- st_transform(applicant_zips_points_sf, st_crs(chicago_boundary))
property_zips_points_sf <- st_transform(property_zips_points_sf, st_crs(chicago_boundary))

# Clip points so they are limited to cook county
applicant_zips_points_sf <- st_intersection(applicant_zips_points_sf, chicago_boundary)
property_zips_points_sf <- st_intersection(property_zips_points_sf, chicago_boundary)

# Extract coordinates first from applicant points
applicant_zips_points_sf <- applicant_zips_points_sf %>%
  mutate(
    INTPTLON10 = st_coordinates(.)[,1], 
    INTPTLAT10 = st_coordinates(.)[,2]
  ) %>%
  st_drop_geometry()  # Drop geometry before joining

# Extract coordinates first from property points & shift west
property_zips_points_sf <- property_zips_points_sf %>%
  mutate(
    INTPTLON10 = st_coordinates(.)[,1] - 0.0068,  # Shift west
    INTPTLAT10 = st_coordinates(.)[,2]
  ) %>%
  st_drop_geometry()  # Drop geometry before joining

# Ensure property_zips_points_sf column names are unique BEFORE join
property_zips_points_sf <- property_zips_points_sf %>%
  rename(
    INTPTLON10_prop = INTPTLON10,  # Rename longitude for property zips
    INTPTLAT10_prop = INTPTLAT10   # Rename latitude for property zips
  )

# Perform inner join
apptoprop_df <- inner_join(
  applicant_zips_points_sf, property_zips_points_sf, 
  by = "Record.ID"
)

# Filter dataframe to only applications for properties with more than 5 applications
apptoprop_df <- apptoprop_df %>%
  filter(ZCTA5CE10.y == 60612)


line_geometries <- st_sfc(
  lapply(1:nrow(apptoprop_df), function(i) {
    st_linestring(rbind(
      c(apptoprop_df$INTPTLON10[i], apptoprop_df$INTPTLAT10[i]),  # Start point
      c(apptoprop_df$INTPTLON10_prop[i], apptoprop_df$INTPTLAT10_prop[i]) # End point
    ))
  }),
  crs = 4269  # Ensure correct CRS (change if needed)
)

# Convert the dataframe into an sf object with the geometry column
apptoprop_sf <- st_sf(apptoprop_df, geometry = line_geometries)

applicant_zips_points_sf <- applicant_zips_points_sf %>%
   filter(Property_ZipCode == 60612)

property_zips_points_sf <- property_zips_points_sf %>%
   filter(ZCTA5CE10 == 60612)

external_zips <- property_zips_points_sf %>%
  group_by(applicant_ZIP) %>%
  summarise(Count = n())


# Ensure applicant points are sf objects
applicant_zips_points_sf <- st_as_sf(applicant_zips_points_sf, 
                                     coords = c("INTPTLON10", "INTPTLAT10"), 
                                     crs = 4269)  # Adjust CRS as needed

# Ensure property points are sf objects
property_zips_points_sf <- st_as_sf(property_zips_points_sf, 
                               coords = c("INTPTLON10_prop", "INTPTLAT10_prop"), 
                               crs = 4269)

apptoprop_sf <- st_transform(apptoprop_sf, crs = 4326)
applicant_zips_points_sf <- st_transform(applicant_zips_points_sf, crs = 4326)
property_zips_points_sf <- st_transform(property_zips_points_sf, crs = 4326)

# Create the leaflet map
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%  # Minimal basemap

  # Add ZIP Code Outlines (Dark Grey)
  addPolygons(data = zip_shapes, 
              color = "darkgrey",  # Outline color
              weight = 1,  # Line thickness
              fill = FALSE,  # No fill
              opacity = 1) %>%
  
  
  # Add Lines (Black)
  addPolylines(data = apptoprop_sf, color = "black", weight = 1) %>%
  
  # Add Start Points (Applicant ZIPs) in Blue
  addCircleMarkers(data = applicant_zips_points_sf, 
                   color = "blue", radius = 2) %>%
  
  # Add End Points (Property ZIPs) in Red
  addCircleMarkers(data = property_zips_points_sf, 
                   color = "red", radius = 2) %>%

# Set View to Chicago
  setView(lng = -87.6298, lat = 41.8781, zoom = 11)  # Chicago's coordinates




```

# Exporting Shapefiles for Figure 16

```{r}

fix_field_names <- function(names_vec, max_length = 9) {
  new_names <- character(length(names_vec))
  used_names <- character(0)
  for (i in seq_along(names_vec)) {
    # Start with a truncated base name
    base <- substr(names_vec[i], 1, max_length)
    new_name <- base
    suffix <- 1
    # Ensure the new name is unique
    while(new_name %in% used_names) {
      # Reserve room for underscore and suffix
      avail <- max_length - nchar(as.character(suffix)) - 1 
      new_name <- paste0(substr(base, 1, avail), "_", suffix)
      suffix <- suffix + 1
    }
    new_names[i] <- new_name
    used_names <- c(used_names, new_name)
  }
  new_names
}

# Fix field names for each sf object
names(applicant_zips_points_sf) <- fix_field_names(names(applicant_zips_points_sf), max_length = 9)
names(property_zips_points_sf)  <- fix_field_names(names(property_zips_points_sf), max_length = 9)

# (Optional) Check that the names are unique
if(any(duplicated(names(applicant_zips_points_sf)))) {
  stop("There are still duplicates in applicant_zips_points_sf!")
}
if(any(duplicated(names(property_zips_points_sf)))) {
  stop("There are still duplicates in property_zips_points_sf!")
}

# Write to shapefiles
# st_write(applicant_zips_points_sf, "Figure_16_Applicant_Dots.shp", driver = "ESRI Shapefile")
# st_write(property_zips_points_sf,  "Figure_16_Property_Dots.shp", driver = "ESRI Shapefile")

```




# Percent of applications from within the same ZIP code for non-individual entities

```{r}

data_clean_group <- data_clean %>%
  filter(Entity_Type_Group == "All Others")

# For zip codes that are the same for applicant and property, summarise applications (rows) by zip code and count the number of times that each zip appears in a new column called "local_count"
matched_zips <- data_clean_group %>%
  filter(applicant_ZIP == Property_ZipCode) %>%
  group_by(applicant_ZIP) %>%
  summarise(local_count = n())

# For ALL zip codes, summarise applications (rows) by zip code and count the number of times that each zip appears in a new column called "time_count"
total_zips <- data_clean %>%
  filter(applicant_ZIP == Property_ZipCode) %>%
  group_by(applicant_ZIP) %>%
  summarise(total_count = n())

joined_zips <- inner_join(total_zips,matched_zips, by = "applicant_ZIP")

# Calculate the percentage of local applicants
joined_zips <- joined_zips %>%
  mutate(percent_local = (local_count / total_count) * 100)

percent_local_zips <- left_join(zips_chi_join, joined_zips, by = c("ZCTA5CE10" = "applicant_ZIP"))

percent_local_zips <- percent_local_zips %>%
  mutate(percent_local = replace_na(percent_local, 0))

tm_shape(percent_local_zips) +
  tm_polygons(fill = "percent_local", 
              fill.scale = tm_scale_intervals(n = 5, style = "equal", values = "brewer.blues")) +
  tm_layout(legend.outside = TRUE, frame = FALSE) +
  tm_title("Percent of Applications from within ZIP Code for non-individual entities")


```

# Percent of applications going to different ZIP Codes

```{r}

zip_shapes 

# For zip codes that are different for applicant and property, summarise applications (rows) by zip code and count the number of times that each zip # # appears in a new column called "non_local_count"
non_matched_zips <- data_clean %>%
  filter(applicant_ZIP != Property_ZipCode) %>%
  group_by(applicant_ZIP) %>%
  summarise(non_local_count = n())

# For ALL zip codes, summarise applications (rows) by zip code and count the number of times that each zip appears in a new column called "total_count"
total_zips <- data_clean %>%
  group_by(applicant_ZIP) %>%
  summarise(total_count = n())

joined_zips <- inner_join(total_zips,non_matched_zips, by = "applicant_ZIP")

# Calculate the percentage of non-local applicants
joined_zips <- joined_zips %>%
  mutate(percent_non_local = (non_local_count / total_count) * 100)

percent_non_local_zips <- left_join(zips_chi_join, joined_zips, by = c("ZCTA5CE10" = "applicant_ZIP"))

percent_non_local_zips <- percent_non_local_zips %>%
  mutate(percent_non_local = replace_na(percent_non_local, 0))

tm_shape(percent_non_local_zips) +
  tm_polygons(fill = "percent_non_local", 
              fill.scale = tm_scale_intervals(n = 5, style = "jenks", values = "brewer.blues")) +
  tm_layout(legend.outside = TRUE, frame = FALSE) +
  tm_title("Percent of Applications to properties in different ZIP Code")

```

# Percent of applications going to different ZIP Codes

```{r}

zip_shapes 

# For zip codes that are different for applicant and property, summarise applications (rows) by zip code and count the number of times that each zip # # appears in a new column called "non_local_count"
non_matched_zips <- data_clean %>%
  filter(applicant_ZIP != Property_ZipCode) %>%
  group_by(applicant_ZIP) %>%
  summarise(non_local_count = n())

# For ALL zip codes, summarise applications (rows) by zip code and count the number of times that each zip appears in a new column called "total_count"
total_zips <- data_clean %>%
  group_by(applicant_ZIP) %>%
  summarise(total_count = n())

invest_zips <- data_clean %>%
  filter(Property_ZipCode == "60612")

joined_zips <- inner_join(total_zips,non_matched_zips, by = "applicant_ZIP")

# Calculate the percentage of non-local applicants
joined_zips <- joined_zips %>%
  mutate(percent_non_local = (non_local_count / total_count) * 100)

percent_non_local_zips <- left_join(zips_chi_join, joined_zips, by = c("ZCTA5CE10" = "applicant_ZIP"))

percent_non_local_zips <- percent_non_local_zips %>%
  mutate(non_local_count = replace_na(non_local_count, 0))

tm_shape(percent_non_local_zips) +
  tm_polygons(fill = "non_local_count", 
              fill.scale = tm_scale_intervals(n = 5, style = "jenks", values = "brewer.blues")) +
  tm_layout(legend.outside = TRUE, frame = FALSE) +
  tm_title("Percent of Applications from outside property ZIP Code")

```

# Count of applications coming from outside ZIP Codes

```{r}

zip_shapes 

# For zip codes that are different for applicant and property, summarise applications (rows) by zip code and count the number of times that each zip # # appears in a new column called "non_local_count"
non_matched_zips <- data_clean %>%
  filter(applicant_ZIP != Property_ZipCode) %>%
  group_by(Property_ZipCode) %>%
  summarise(outside_app_count = n())

# Convert Property_ZIPCode to character
non_matched_zips <- non_matched_zips %>%
  mutate(Property_ZipCode = as.character(Property_ZipCode))

count_non_local_apps <- left_join(zips_chi_join, non_matched_zips, by = c("ZCTA5CE10" = "Property_ZipCode"))

count_non_local_apps <- count_non_local_apps %>%
  mutate(outside_app_count = replace_na(outside_app_count, 0))

tm_shape(count_non_local_apps) +
  tm_polygons(fill = "outside_app_count", 
              fill.scale = tm_scale_intervals(n = 5, style = "jenks", values = "brewer.blues")) +
  tm_layout(legend.outside = TRUE, frame = FALSE) +
  tm_title("Count of Applications from outside property ZIP Code")


```

